---
title: "Export Pretty Tables"
execute: 
  message: false
  warning: false
  freeze: auto
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))

supp_dir <- here("data", "supplementary_tables", "csvs")
```

```{r}
immune_hi.v.lo_df <- readRDS(file.path(results_dir, "immune_hi.v.lo.rds"))
bulk_clin_metadata <- read_tsv(file.path(data_dir, "bulk", "bulk-clin-metadata.tsv")) %>% 
  # filter(is.na(excluded_clin) & is.na(excluded_rna)) %>% 
  # drop_na(sample_RNA) %>% 
  select(-immune_hi.v.lo, -immune_zscore, -immune_hi.v.lo_zscore) %>% 
  left_join(immune_hi.v.lo_df, by = "sample_RNA") %>% 
  dplyr::select(-immune_hi.med.lo_zscore, -immune_hi.v.lo_zscore, -kcluster, 
                -group)
immune_signatures <- read_tsv(file.path(results_dir, "immune_signatures.tsv"))$signatures

gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
ssgsea_df <- read_tsv(file.path(results_dir, "ssgsea_df.tsv"))
ssgsea_tidy <- ssgsea_df %>% 
  pivot_longer(cols = 2:ncol(.), 
               names_to = "sample_RNA", 
               values_to = "score") %>% 
  group_by(gs_name) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup()

dge_immune_df <- read_tsv(file.path(results_dir, "dge_treatment_immune_df.tsv"))
gsea_immune_df <- read_tsv(file.path(results_dir, "gsea_R.v.NR_by_treatment_immune.tsv"))

dge_df <- read_tsv(file.path(results_dir, "dge_treatment.tsv"))
gsea_df <- read_tsv(file.path(results_dir, "gsea_treatment.tsv"))
```

# Tables

1.  Full details of the meta-cohort
2.  RNA metadata
3.  Batch-corrected TPMs
4.  DGE, immune stratified
5.  GSEA, immune stratified
6.  estimated effect of hypoxia per cell type
7.  CellPhoneDB results

# Supp 1

Full details of the meta-cohort

```{r}
bulk_clin_metadata %>% colnames() 
```

```{r}
bulk_clin_metadata %>% 
  mutate_at(c("age", "os_days", "pfs_days",
              "censor_os_alive_1_dead_0", 'censor_pfs_event_1_progressed_0'),
            as.integer) %>%  
  dplyr::rename(response_noSD = response) %>% 
  dplyr::rename(response = response_splitSD) %>% 
  dplyr::select(-excluded_wes, -excluded_rna, -sequencing_modality, 
                -immune_hi.v.lo, -immune_zscore) %>% 
  write_excel_csv(file.path(supp_dir, "SupplementaryTable1.csv"))
```

```{r}
bulk_clin_metadata %>% 
  group_by(sex) %>% 
  count()
```

```{r}
bulk_clin_metadata %>% 
  filter(is.na(excluded_clin) & is.na(excluded_rna)) %>% 
  drop_na(sample_RNA) %>% 
  group_by(sex) %>% 
  count()
```

# Supp 2

```{r}
star_df <- readRDS(here("data", "bulk", "star_df.rds"))
```

```{r}
read_tsv(here("data", "bulk", "bulk-rna-metadata.tsv")) %>% 
  dplyr::select(sample_RNA, excluded_rna, sequencing_modality) %>% 
  left_join(immune_hi.v.lo_df, by = "sample_RNA") %>% 
  dplyr::select(-immune_hi.v.lo_zscore, -immune_hi.med.lo_zscore, -kcluster) %>% 
  left_join(star_df, by = c("sample_RNA" = "sample")) %>% 
  write_excel_csv(file.path(supp_dir, "SupplementaryTable2.csv"))
```

# Supp 3

Batch-corrected TPMs (ComBat-seq), genes as rows and samples as columns.

```{r}
tpm_df_combatseq <- readRDS(file.path(results_dir, "tpm_df_combatseq.rds"))
```

```{r}
tpm_df_combatseq %>%
  ungroup() %>%
  pivot_longer(cols = -sample_RNA, names_to = "gene", values_to = "tpm") %>%
  pivot_wider(names_from = sample_RNA, values_from = tpm) %>%
  write_excel_csv(file.path(supp_dir, "SupplementaryTable3.csv"))
```

# Supp 4

DGE, immune stratified

```{r}
dge_immune_df %>% 
  colnames()
```

```{r}
dge_immune_df_cleaned <- dge_immune_df %>% 
  dplyr::select(treatment, immune_hi.v.lo, gene, p.value, p_BH, 
                median_nr, median_r, med_fc, 
                mean_cr, mean_r, avg_fc, 
                signed_neglog10_p, 
                n_R, n_NR)
write_excel_csv(dge_immune_df_cleaned, 
          file.path(supp_dir, "SupplementaryTable4.csv"))
```

# Supp 5

```{r}
gsea_immune_df %>% colnames()
```

```{r}
write_excel_csv(gsea_immune_df, 
          file.path(supp_dir, "SupplementaryTable5.csv"))
```

# Supp 6

```{r}
pb_ols <- readRDS(here(
  "results",
  "sc-analysis-hypoxia-pb-gene-expr-modeling-simplified.rds"
))  %>%
  group_by(cell_type) %>%
  mutate(padj = p.adjust(p.value, method = "BH")) %>%
  ungroup()
```

```{r}
pb_ols %>% 
  write_excel_csv(file.path(supp_dir, "SupplementaryTable6.csv"))
```

# Supp 7

```{r}
cpdb_tidy_stromal.imm <- 
  readRDS(file.path(results_dir, "cpdb_tidy_stromal.imm.rds"))
```

```{r}
cpdb_wide <- cpdb_tidy_stromal.imm %>% 
  pivot_wider(id_cols = c(id, id_cp_interaction, interacting_pair, 
                          partner_a, partner_b, 
                          gene_a, gene_b, secreted, 
                          receptor_a, receptor_b, 
                          annotation_strategy, 
                          is_integrin), 
              names_from = name, 
              values_from = c(value.mean, value.pval))
colnames(cpdb_wide) <- str_replace_all(colnames(cpdb_wide), fixed("Mac/mono"), "TAM")
```


```{r}
cpdb_wide %>% 
  write_excel_csv(file.path(supp_dir, "SupplementaryTable7.csv"))
```

# Additional tables

Immune signatures + immune score modeling

```{r}
predict_df <- bulk_clin_metadata %>% 
  filter(is.na(excluded_clin) & is.na(excluded_rna)) %>%
  drop_na(sample_RNA) %>%
  select(sample_RNA, immune_zscore, immune_hi.v.lo, response_splitSD, treatment, cohort) %>% 
  drop_na(response_splitSD) %>% 
  mutate(immune_hi.v.lo = factor(immune_hi.v.lo, levels = c("immune low", "immune high"), ordered = TRUE)) %>% 
  mutate(response = case_when(
    response_splitSD == "R" ~ 1, 
    response_splitSD == "NR" ~ 0
  )) %>% 
  mutate(treatment = factor(
    treatment,
    levels = c("aPD1", "aCTLA4", "combo"),
    ordered = FALSE
  )) %>% 
  column_to_rownames("sample_RNA")
```

```{r}
models <- list(
  "response ~ 1",
  "response ~ immune_zscore", 
  "response ~ treatment", 
  "response ~ treatment + immune_zscore", 
  "response ~ treatment * immune_zscore"
) %>% set_names()
```

```{r}
glm_out_list <- models %>% 
  map(~glm(formula(..1), data = predict_df, family = "binomial"))
aic_df <- glm_out_list %>% 
  map_dfr(~tibble(AIC = ..1$aic), .id = "variable") %>% 
  arrange(AIC) %>% 
  mutate(delta_AIC = AIC - min(AIC)) 
print(aic_df)
```

```{r}
immunescore_df <- broom::tidy(glm_out_list$`response ~ treatment * immune_zscore`)
print(immunescore_df)
```

```{r}
models <- list(
  "response ~ 1",
  "response ~ immune_hi.v.lo", 
  "response ~ treatment", 
  "response ~ treatment + immune_hi.v.lo", 
  "response ~ treatment * immune_hi.v.lo"
) %>% set_names()
```

```{r}
glm_out_list <- models %>% 
  map(~glm(formula(..1), data = predict_df, family = "binomial"))
aic_df <- glm_out_list %>% 
  map_dfr(~tibble(AIC = ..1$aic), .id = "variable") %>% 
  arrange(AIC) %>% 
  mutate(delta_AIC = AIC - min(AIC)) 
print(aic_df)
```

```{r}
immune_df <- broom::tidy(glm_out_list$`response ~ treatment * immune_hi.v.lo`)
print(immune_df)
```

```{r}
immune_df <- broom::tidy(glm_out_list$`response ~ treatment`)
print(immune_df)
```

