---
title: "Jerby-arnon analysis"
execute: 
  freeze: auto
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
source(here("lib", "single_cell", "helper_Seurat.R"))

# document specific packages: 
library(Seurat)
library(UCell)

file_cache_dir <- file.path(cache_dir, "external-sc-jerby-arnon")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

# Set up

## Read in seurat objects 

```{r}
so <- readRDS(file.path(data_dir, "scRNAseq", "so_jerbyarnon.rds"))
```

## Set up some data frames

```{r}
metadata_df <- get_metadata_from_so(so) %>% 
  mutate(cell.type.renamed = case_when(
    cell.type == "?" ~ "Unknown", 
    cell.type == "B.cell" ~ "B cell", 
    cell.type == "CAF" ~ "Fibroblasts", 
    cell.type == "Endo." ~ "Endothelial cells", 
    cell.type == "Macrophage" ~ "TAM", 
    cell.type == "Mal" ~ "Tumor", 
    cell.type == "T.CD8" ~ "CD8+ T", 
    cell.type %in% c("T.CD4", "T.cell") ~ "T cell", 
    TRUE ~ cell.type
  ))

sample_metadata_df <- so@meta.data %>% 
  as_tibble(rownames = "cell") %>% 
  group_by(Sample, Treatment.group, Lesion.type, Treatment) %>% 
  summarise(ncells = n())
```

# UMAP

```{r}
plot_umap(metadata_df, color = "cell.type.renamed", pointsize = 2.1) + 
  scale_color_manual(values = c(palette_general_cell_type[-6],
                                palette_specific_immune,
                                c("T cell" = "#A6D3A0")))
plot_umap(metadata_df, color = "Treatment.group", pointsize = 2.1)
plot_umap(metadata_df, color = "Lesion.type", pointsize = 2.1) 
```

```{r, fig.asp = 1/4, fig.width = 12}
plot_umap(metadata_df, color = "Site", pointsize = 2.1) 
```

```{r}

metadata_df %>% 
  mutate(Site_grouped = case_when(
    str_detect(Site, "ymph node|Spleen")  ~ "LN/Spleen", 
    str_detect(Site, "Skin|Cut|cut|Acral") ~ "Skin", 
    TRUE ~ "other"
  )) %>% 
  plot_umap(color = "Site_grouped", pointsize = 2.1)
```


# UCell signature scoring

## Hallmarks and Bagaev

```{r}
signature_df <- read_excel(
  file.path(data_dir, "signatures", "1-s2.0-S1535610821002221-mmc2.xlsx"), 
  sheet = "Gene Description", 
  skip = 1
) %>% mutate(Gene = str_replace(Gene, "-", "."))

signature_list <- signature_df %>% 
  named_group_split(`Gene signature`) %>% 
  map(~str_replace(.$Gene, "\\.", "-")) 
```

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
select_gs <- c(
  "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
  "HALLMARK_INTERFERON_ALHPA_RESPONSE", 
  "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", 
  "HALLMARK_ANGIOGENESIS", 
  "HALLMARK_HYPOXIA"
)
select_gs_vec <- gs_vec[select_gs]
```

```{r}
ucell_h <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_stromal_scores.rds"), 
  function() {
    ScoreSignatures_UCell(so@assays$RNA@data,
                          features = c(signature_list, select_gs_vec))
  }
)
```

* CXCL8,IFNA2,VSIR,TRAC,TRBC1,TRBC2,CMTR1,MARCHF1,WARS1,CCN1,CCN2,COLGALT1,P3H1,ACKR3,CAVIN1,CAVIN3,CCN5,ERO1A,LARGE1,NOCT

## Benci et al. 

```{r}
ifng_gs <- read_csv(file.path(data_dir, "signatures", "ifng_gs.csv"))$IFNG.GS
isg_rs <- read_csv(file.path(data_dir, "signatures", "isg_rs.csv"))$ISG.RS

benci_list <- list(
  ISG.RS = isg_rs, 
  IFNG.GS = ifng_gs
)

```

```{r}
ucell_benci <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_stromal_benci.rds"), 
  function() {
    ScoreSignatures_UCell(so@assays$RNA@data,
                          features = c(benci_list))
  }
)
```

Warning: The following genes were not found and will be
                        imputed to exp=0:
* PRIC285

# Characteristics of the data

```{r}
metadata_df %>% 
  group_by(Sample) %>% 
  count() %>% 
  left_join(sample_metadata_df) %>% 
  ggplot() + 
  aes(fct_reorder(Sample, n), n, color = Treatment.group) + 
  geom_point(size = 2) + 
  geom_segment(aes(xend = Sample, yend = 0)) + 
  rotate_x_text() +
  remove_x_spine()
```


# Immune stratification

```{r}
prop_immune_df <- so@meta.data %>% 
  as_tibble(rownames = "cell") %>% 
  filter(cell.type.general != "Unknown") %>% 
  group_by(Sample, cell.type.general) %>% 
  count() %>% 
  group_by(Sample) %>% 
  mutate(total = sum(n), proportion = n / total) %>% 
  pivot_wider(id_cols = Sample, names_from = cell.type.general, 
              values_from = c(n, proportion), values_fill = 0)
```

So there are 12 samples that are untreated, from metastases. 

# Hypoxia stratification

## Hypoxia score per cell type

```{r, fig.asp = 1.618, fig.width = 3}
ucell_h %>% 
  as_tibble(rownames = "cell") %>% 
  left_join(metadata_df, by = "cell") %>% 
  filter(Lesion.type == "metastasis") %>% 
  filter(Treatment.group == "Untreated") %>% 
  filter(cell.type.general != "Unknown") %>% 
  ggplot() + 
  aes(cell.type.general, HALLMARK_HYPOXIA_UCell, 
      fill = cell.type.general) + 
  geom_violin(scale = "width", 
              draw_quantiles = 0.5, 
              color = "white", 
              size = 0.1) + 
  rotate_x_text() + 
  remove_x_spine() + 
  scale_fill_manual(values = palette_compartment) + 
  theme(legend.position = "none")
```

```{r}
scaled_hypoxia_ucell_df <- ucell_h %>% 
  as_tibble(rownames = "cell") %>% 
  left_join(metadata_df, by = "cell") %>% 
  filter(Lesion.type == "metastasis") %>% 
  filter(Treatment.group == "Untreated") %>% 
  filter(cell.type.general != "Unknown") %>% 
  group_by(cell.type.general) %>% 
  mutate(hypoxia_scaled = as.vector(scale(HALLMARK_HYPOXIA_UCell))) 
scaled_hypoxia_ucell_df %>% 
  ggplot() + 
  aes(cell.type.general, hypoxia_scaled, 
      fill = cell.type.general) + 
  geom_violin(scale = "width", 
              draw_quantiles = 0.5, 
              color = "white", 
              size = 0.1) + 
  rotate_x_text() + 
  remove_x_spine() + 
  scale_fill_manual(values = palette_compartment) + 
  theme(legend.position = "none")
```

```{r}
scaled_hypoxia_ucell_selected_df <- scaled_hypoxia_ucell_df %>% 
  ungroup() %>% 
  select(cell, hypoxia_scaled)
```

## Sample level

```{r}
hypoxia_all_df <- ucell_h %>% 
  as_tibble(rownames = "cell") %>% 
  left_join(metadata_df, by = "cell") %>% 
  filter(cell.type.general != "Unknown") %>% 
  group_by(Sample) %>% 
  summarise(mean_hypoxia = mean(HALLMARK_HYPOXIA_UCell), 
            n = n()) %>% 
  mutate(mean_hypoxia = as.vector(scale(mean_hypoxia)), 
         sd_hypoxia = sd(mean_hypoxia)) %>% 
  left_join(sample_metadata_df, by = "Sample") %>% 
  # mutate(hypoxia_hi.v.lo = ifelse(
  #   mean_hypoxia > median(mean_hypoxia), "hypoxic", "non-hypoxic"
  # ))
  mutate(hypoxia_hi.v.lo = case_when(
    mean_hypoxia > quantile(mean_hypoxia, 0.667) ~ "hypoxic", 
    mean_hypoxia > quantile(mean_hypoxia, 0.333) ~ "mid", 
    TRUE ~ "non-hypoxia"
  ))
hypoxia_all_df %>% 
  ggplot() + 
  aes(fct_reorder(Sample, mean_hypoxia), mean_hypoxia, 
      color = hypoxia_hi.v.lo) + 
  geom_point(size = 1) + 
  geom_segment(aes(
    xend = Sample,
    y = mean_hypoxia - sd_hypoxia,
    yend = mean_hypoxia + sd_hypoxia
  )) +
  rotate_x_text() +
  remove_x_spine() + 
  labs(x = "Sample", y = "mean hypoxia score")
```

```{r}
hypoxia_untreated_met_df <- ucell_h %>% 
  as_tibble(rownames = "cell") %>% 
  left_join(metadata_df, by = "cell") %>% 
  filter(Lesion.type == "metastasis") %>% 
  filter(Treatment.group == "Untreated") %>% 
  filter(cell.type.general != "Unknown") %>% 
  left_join(scaled_hypoxia_ucell_selected_df, by = "cell") %>% 
  group_by(Sample) %>% 
  summarise(mean_hypoxia = mean(hypoxia_scaled), 
            n = n()) %>% 
  mutate(mean_hypoxia = as.vector(scale(mean_hypoxia)), 
         sd_hypoxia = sd(mean_hypoxia)) %>% 
  left_join(sample_metadata_df, by = "Sample") %>% 
  # mutate(hypoxia_hi.v.lo = ifelse(
  #   mean_hypoxia > median(mean_hypoxia), "hypoxic", "non-hypoxic"
  # ))
  mutate(hypoxia_hi.v.lo = case_when(
    mean_hypoxia > quantile(mean_hypoxia, 0.667) ~ "high hypoxia", 
    mean_hypoxia > quantile(mean_hypoxia, 0.333) ~ "mid hypoxia", 
    TRUE ~ "low hypoxia"
  ))
hypoxia_untreated_met_df %>% 
  ggplot() + 
  aes(fct_reorder(Sample, mean_hypoxia), mean_hypoxia, 
      color = hypoxia_hi.v.lo) + 
  geom_point(size = 1) + 
  geom_segment(aes(
    xend = Sample,
    y = mean_hypoxia - sd_hypoxia,
    yend = mean_hypoxia + sd_hypoxia
  )) +
  rotate_x_text() +
  remove_x_spine()
```

## Correlation 

```{r}
hypoxia_compartment_df <- ucell_h %>% 
  as_tibble(rownames = "cell") %>% 
  left_join(metadata_df, by = "cell") %>% 
  filter(Lesion.type == "metastasis") %>%
  filter(Treatment.group == "Untreated") %>%
  left_join(scaled_hypoxia_ucell_selected_df, by = "cell") %>% 
  group_by(Sample, cell.type.general) %>% 
  summarise(mean_hypoxia = mean(hypoxia_scaled)) %>% 
  left_join(sample_metadata_df, by = "Sample") 

hypoxia_compartment_wide <- hypoxia_compartment_df %>% 
  filter(cell.type.general != "Unknown") %>% 
  pivot_wider(id_cols = Sample, 
              names_from = cell.type.general, 
              values_from = mean_hypoxia)

```

```{r, fig.asp = 1, fig.width = 3}
hypoxia_compartment_wide %>% 
  ggplot() + 
  aes(Immune, Tumor) + 
  geom_point(size = 4)
hypoxia_compartment_wide %>% 
  ggplot() + 
  aes(Stromal, Tumor) + 
  geom_point(size = 4)
hypoxia_compartment_wide %>% 
  ggplot() + 
  aes(Stromal, Immune) + 
  geom_point(size = 4)
```



```{r, fig.width = 1.3, fig.height = 1.5}
cor_mtx <- hypoxia_compartment_wide %>% 
  drop_na(Stromal, Tumor) %>% 
  column_to_rownames("Sample") %>%
  cor(method = "pearson") 
h_cor <- cor_mtx %>%
  Heatmap(
    col = colorRamp2(breaks = c(0, 1), colors = c("black", "white")),
    name = "Pearson cor.",
    row_names_gp = gpar(fontsize = 7),
    column_names_gp = gpar(fontsize = 7),
    row_dend_width = unit(2, "mm"),
    row_dend_gp = gpar(lwd = 0.2), 
    column_dend_height = unit(2, "mm"), 
    column_dend_gp = gpar(lwd = 0.2), 
    heatmap_legend_param = list(
      title_gp = gpar(fontsize = 7),
      labels_gp = gpar(fontsize = 7),
      legend_width = unit(10, "mm"),
      grid_height = unit(1, "mm"), 
      title_position = "leftcenter", 
      legend_direction = "horizontal"
    ), 
    cell_fun = function(j, i, x, y, w, h, col) {
      grid.text(format(cor_mtx[i, j], digits = 2), x, y, 
                gp = gpar(fontsize = 5))
    }
  )
draw(h_cor, heatmap_legend_side = "bottom")
```


## subset so object

```{r}
so_untreated_met_subset <- subset(so %>% set_idents("Lesion.type"), 
                                  idents = "metastasis")
so_untreated_met_subset <- subset(so_untreated_met_subset %>%
                                    set_idents("Treatment.group"), 
                                  idents = "Untreated")

select_hypoxia_untreated_met_df <- hypoxia_untreated_met_df %>% 
  select(Sample, hypoxia_hi.v.lo) 
metadata_to_add <- so_untreated_met_subset@meta.data %>% 
  select(Sample, cell.type) %>% 
  as_tibble(rownames = "cell") %>% 
  left_join(select_hypoxia_untreated_met_df, by = "Sample") %>%  
  mutate(cell.type.renamed = case_when(
    cell.type == "?" ~ "Unknown", 
    cell.type == "B.cell" ~ "B cell", 
    cell.type == "CAF" ~ "Fibroblasts", 
    cell.type == "Endo." ~ "Endothelial cells", 
    cell.type == "Macrophage" ~ "TAM", 
    cell.type == "Mal" ~ "Tumor", 
    cell.type == "T.CD8" ~ "CD8+ T", 
    cell.type %in% c("T.CD4", "T.cell") ~ "T cell", 
    TRUE ~ cell.type
  )) %>% 
  column_to_rownames("cell") %>% 
  select(-Sample, -cell.type)
so_untreated_met_subset <- AddMetaData(so_untreated_met_subset, metadata = metadata_to_add)
```


# Cell types proportions by hypoxia

```{r}
immsubtype_prop_df <- so_untreated_met_subset@meta.data %>% 
  as_tibble(rownames = "cell") %>% 
  filter(cell.type.general != "Unknown") %>% 
  group_by(Sample, cell.type.renamed) %>% 
  count() %>% 
  group_by(Sample) %>% 
  mutate(total = sum(n), proportion = n / total)
# %>% 
#   pivot_wider(id_cols = Sample, names_from = cell.type.renamed, 
              # values_from = c(n, proportion), values_fill = 0)
```


```{r, fig.asp = 1/4, fig.width = 12}
immsubtype_prop_df %>% 
  left_join(hypoxia_untreated_met_df, by = "Sample") %>% 
  ggplot() + 
  aes(hypoxia_hi.v.lo, proportion) + 
  geom_boxplot() + 
  facet_wrap(~cell.type.renamed, ncol = 8, scale = "free_y") + 
  rotate_x_text()
```


# Characterizing Hypoxia 

## Hypoxia vs immune stratification

```{r}
hypoxia_untreated_met_df %>%
  left_join(prop_immune_df, by = "Sample") %>%
  ggplot() +
  aes(proportion_Immune, mean_hypoxia) +
  geom_point(size = 3) +
  stat_smooth(method = "lm") +
  stat_cor() +
  scale_x_continuous(limits = c(0, 1))
```

## Expression of hypoxia-related transcription factors

```{r, fig.asp = 1/4, fig.width = 10}
hypoxia_genes <- c("EPAS1", "HIF1A", "VEGFA", "FLT1", "KDR")
hypoxia_gene_df <- get_gene_expr_from_so(so_untreated_met_subset, gene_vec = hypoxia_genes)

metadata_df %>% 
  filter(cell.type.general != "Unknown") %>% 
  left_join(hypoxia_gene_df, by = "cell") %>%
  pivot_longer(cols = hypoxia_genes) %>% 
  group_by(Sample, cell.type.renamed, name) %>% 
  summarise(mean_value = mean(value)) %>% 
  drop_na(mean_value) %>% 
  left_join(hypoxia_untreated_met_df, by = "Sample") %>% 
  ggplot() + 
  aes(cell.type.renamed, mean_value) + 
  geom_col(aes(y = mean_mean_value, fill = cell.type.renamed), 
           data = ~..1 %>% 
             group_by(cell.type.renamed, name) %>% 
             summarise(mean_mean_value = mean(mean_value))) +
  geom_jitter(width = 0.2, size = 1) + 
  scale_fill_manual(values = c(palette_general_cell_type, 
                               palette_specific_immune)) + 
  facet_wrap( ~ name, ncol = 5, scale = "free_x") +
  theme(legend.position = "none") + 
  coord_flip() 
```

# pb

```{r}
make_specific_pb_hypoxia <- function(so, cluster_str, ident_str) {
  so_subset <- subset(so %>% set_idents(cluster_str),
                      ident = ident_str)
  so_subset@meta.data$Sample <-
    factor(so_subset@meta.data$Sample)
  pb <- make_pseudobulk(
    so_subset,
    split_var = NULL,
    sample_var = "Sample",
    vars = c("hypoxia_hi.v.lo")
  )
  return(pb)
}

pb_list <- run_or_readRDS(
  file.path(file_cache_dir, "pb.rds"), 
  function() {
    pb_imm_list <-
      so_untreated_met_subset@meta.data$cell.type.renamed %>% unique() %>%
      set_names() %>%
      .[-3] %>% # drop ?
      map( ~ make_specific_pb_hypoxia(so_untreated_met_subset, "cell.type.renamed", ..1))
  }
)
```


# Setting up modeling

## Attempt: y=log1p(CPM), OLS

```{r}
pb_df <- pb_list %>% 
  map_dfr(as_tibble, rownames = "sample_ID", .id = "cell_type") %>% 
  separate(sample_ID, sep = "_", into = c("sample_ID", "hypoxia_hi.v.lo"))

pb_tidy <- pb_df %>% 
  pivot_longer(cols = 4:ncol(.), 
               names_to = "gene", 
               values_to = "count") %>% 
  mutate(count = ifelse(is.na(count), 0, count)) %>% 
  group_by(sample_ID, cell_type) %>% 
  mutate(total_counts = sum(count), CPM = count / total_counts * 1e6) %>% 
  ungroup() %>% 
  mutate(log1p_CPM = log10(CPM + 1)) %>% 
  left_join(select(hypoxia_untreated_met_df, Sample, mean_hypoxia), 
            by = c("sample_ID" = "Sample"))
```

```{r, eval = FALSE}
saveRDS(pb_tidy, here("results", "jerby-arnon_sc_pb_tidy.rds"))
```

### Model

For each cell type: 

```
log1p_CPM ~ hypoxia
```

And we'll only consider genes with abs(log2FC) > 0.1. 

```{r}
test_genes_in_cell_type <- function(my_pb_tidy, thresh = 0.1) {
  log2fc_df <- my_pb_tidy %>% 
    group_by(gene, hypoxia_hi.v.lo) %>% 
    summarise(mean_log1p_CPM = mean(log1p_CPM)) %>% 
    pivot_wider(id_cols = gene, 
                names_from = hypoxia_hi.v.lo, 
                values_from = mean_log1p_CPM) %>% 
    mutate(log2FC_high_over_low = log2(`high hypoxia` / `low hypoxia`))
  
  select_genes <- log2fc_df %>% 
    filter(abs(log2FC_high_over_low) > thresh) %>% 
    .$gene
  
  my_pb_tidy %>% 
    filter(gene %in% select_genes) %>% 
    group_by(gene) %>% 
    summarise(lm_out = broom::tidy(lm(log1p_CPM ~ mean_hypoxia))) %>% 
    unnest(lm_out) %>% 
    left_join(log2fc_df, by = "gene")
}

```

```{r}
pb_tidy_split <- pb_tidy %>% 
  named_group_split(cell_type)

model_out <- run_or_readRDS(
  file.path(file_cache_dir, "model_out.rds"), 
  function() {
    pb_tidy_split %>% 
      .[-c(3, 4)] %>% # no hypoxia low endothelial cells
      map_dfr(test_genes_in_cell_type, .id = "cell_type")
  }
)
```

Gotta run endothelial cells and fibroblasts separately unfortunately. And it's not great bc there are just no endothelial cells and fibroblasts in the hypoxia low setting

```{r}
model_out_ec <- run_or_readRDS(
  file.path(file_cache_dir, "model_out_ec.rds"), 
  function() {
    # endothelial cells
    pb_tidy %>%
      filter(cell_type == "Endothelial cells") %>%
      group_by(gene) %>%
      summarise(lm_out = broom::tidy(lm(log1p_CPM ~ mean_hypoxia))) %>%
      unnest(lm_out)
  }
)

# fibroblast
model_out_fib <- run_or_readRDS(
  file.path(file_cache_dir, "model_out_fb.rds"), 
  function() {
    pb_tidy %>%
      filter(cell_type == "Fibroblasts") %>%
      group_by(gene) %>%
      summarise(lm_out = broom::tidy(lm(log1p_CPM ~ mean_hypoxia))) %>%
      unnest(lm_out) 
  }
)
```

```{r}
model_out2 <- bind_rows(
  model_out,
  model_out_ec %>% mutate(cell_type = "Endothelial cells"),
  model_out_fib %>% mutate(cell_type = "Fibroblasts")
)
```

## Save

```{r, eval = FALSE}
saveRDS(model_out2, here("results", "jerby-arnon-sc-analysis-hypoxia-pb-gene-expr-modeling-simplified.rds"))
```

# pb ols results 

```{r}
pb_ols <- model_out2
pb_cpm_wide <- pb_tidy %>% 
  pivot_wider(id_cols = c(cell_type, sample_ID), 
              names_from = gene, 
              values_from = log1p_CPM)
mean_hypoxia_df <- hypoxia_untreated_met_df %>% 
  dplyr::rename(sample_ID = Sample)
```


```{r}
plot_pb_ols_ranks <- function(my_pb_ols) {
  my_pb_ols %>% 
    arrange(estimate) %>%
    mutate(rank = row_number()) %>%
    ggplot() +
    aes(rank, estimate) +
    geom_hline(yintercept = 0, linetype = "dotted") +
    geom_line() +
    geom_segment(
      aes(
        x = rank,
        xend = rank,
        y = 0,
        yend = estimate,
        color = category
      ),
      data = ~ ..1 %>% drop_na(category)
    ) +
    geom_point(aes(color = category, shape = p.value < 0.05),
               data = ~ ..1 %>% drop_na(category),
               size = 3) +
    geom_text_repel(
      aes(label = gene, color = category),
      data = ~ ..1 %>% drop_na(category),
      force = 5,
      hjust = 1
    )  + 
    scale_shape_manual(values = c("TRUE" = 19, "FALSE" = 4)) + 
    theme(aspect.ratio = 1)
}
```

## CD8+ T

```{r}
exhaustion_genes <- c("ENTPD1", "CXCL13", "CD101", "HAVCR2", 
                      "LAG3", "PDCD1")#, "CTLA4", "TIGIT")
activation_genes <- c("NFAT5", "TNFSF9")#, "KLRG1", "STAT1", "TAP1")
homing_genes <- c("CCR7")
hypoxia_genes <- c("HIF1A", "EPAS1")

pb_ols %>% 
  ungroup() %>% 
  filter(cell_type == "CD8+ T") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% exhaustion_genes ~ "exhaustion", 
    gene %in% activation_genes ~ "activation", 
    # gene %in% homing_genes ~ "homing", 
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    TRUE ~ NA_character_
  )) %>% 
  plot_pb_ols_ranks() + 
  labs(title = "CD8+ T cells")
```



```{r, fig.asp = 1/4, fig.width = 10}
genes_to_show <- c("ENTPD1", "NFAT5", "CXCL13", "HAVCR2")
exh_cd8_df <- pb_cpm_wide %>% 
  filter(cell_type == "CD8+ T") %>% 
  select(sample_ID, genes_to_show) %>% 
  left_join(mean_hypoxia_df)
exh_cd8_df %>% 
  pivot_longer(cols = genes_to_show) %>% 
  ggplot() + 
  aes(mean_hypoxia, value) + 
  geom_point(size = 2) + 
  stat_smooth(method = "lm", color = "black", se = F) + 
  facet_wrap(~name, scale = "free", ncol = 4) + 
  # theme(legend.position = "top",
  #       legend.direction = "vertical") + 
  scale_color_manual(values = palette_immune_state) + 
  stat_cor(method = "spearman") + 
  labs(y = "normalized expr")
```

# Immunosuppressive TAMs

```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
immunosuppressive_genes <- c("CD163", "SPP1", "CCL2", "CSF1", "CSF1R", "ARG1")
proinflammatory_genes <- c("IL6", "C1QB", "NOS2", "TLR2", "TLR4", "TNF", "IL1B")
adhesion_genes <- c("VCAM1")
cytokines_and_chemokines <- c("IL33", "CXCL6", 
                              # "CXCL10", "CXCL11",
                              "CXCL8", "CXCL2")

pb_ols %>% 
  ungroup() %>% 
  filter(cell_type == "TAM") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    # gene %in% hypoxia_genes ~ "hypoxia response" ,
    gene %in% immunosuppressive_genes ~ "immunosuppressive", 
    gene %in% proinflammatory_genes ~ "proinflammatory", 
    # gene %in% adhesion_genes ~ "adhesion", 
    gene %in% cytokines_and_chemokines ~ "cytokines and chemokines", 
    TRUE ~ NA_character_
  )) %>% 
  plot_pb_ols_ranks() + 
  labs(title = "TAM")
```

```{r, fig.asp = 1/4, fig.width = 10}
genes_to_show <- c("IL1B", "CXCL2", "SPP1")
gene_expr_df <- pb_cpm_wide %>% 
  filter(cell_type == "TAM") %>% 
  select(sample_ID, genes_to_show) %>% 
  left_join(mean_hypoxia_df)
gene_expr_df %>% 
  pivot_longer(cols = genes_to_show) %>% 
  ggplot() + 
  aes(mean_hypoxia, value) + 
  geom_point(size = 2) + 
  stat_smooth(method = "lm", color = "black", se = F) + 
  facet_wrap(~name, scale = "free", ncol = 4) + 
  # theme(legend.position = "top",
  #       legend.direction = "vertical") + 
  scale_color_manual(values = palette_immune_state) + 
  stat_cor(method = "spearman") + 
  labs(y = "normalized expr")
```

# ECs are more angiogenic, less adhesive 

```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
angiogenesis_genes <- c("SERPINE1", "MMP1", "ANGPT2", "PDGFA", "THBS1", "ESM1", "VWF")
cytokines_and_chemokines <- c("IL33", "CXCL6", "CXCL10")
adhesion <- c("EPCAM", "VCAM1")

pb_ols %>% 
  ungroup() %>% 
  filter(cell_type == "Endothelial cells") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    gene %in% angiogenesis_genes ~ "angiogenesis" ,
    gene %in% cytokines_and_chemokines ~ "cytokines and chemokines", 
    gene %in% adhesion ~ "adhesion", 
    TRUE ~ NA_character_
  )) %>% 
  plot_pb_ols_ranks() + 
  labs(title = "Endothelial cells")
```

# Fibroblasts...change?

```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
cytokines_and_chemokines <- c("IL6", "IL33", "CXCL6", "CXCL8")
other <- c("CSF1")

pb_ols %>% 
  ungroup() %>% 
  filter(cell_type == "Fibroblasts") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    gene %in% cytokines_and_chemokines ~ "cytokines and chemokines", 
    gene %in% other ~ "other", 
    TRUE ~ NA_character_
  )) %>% 
  plot_pb_ols_ranks() + 
  labs(title = "Fibroblasts")
```

# Gene signatures

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
hypoxia_vec <- readRDS(file.path(results_dir, "hypoxia_gs_vec.rds"))
prog_vs_term_exh <- readRDS(here("results", "gs_vec_prog.vs.term_exh.rds"))
dc_gs_vec <- readRDS(file.path(results_dir, "dc_gs_vec.rds"))
exh_gs_vec <- readRDS(file.path(results_dir, "exh_gs_vec.rds"))
```

# Run ssGSEA

```{r}
library(GSVA)
```

```{r}
log1p_cpm_mtx <- pb_tidy %>% 
  pivot_wider(id_cols = gene, 
              names_from = c(sample_ID, cell_type), 
              values_from = log1p_CPM, 
              values_fill = 0) %>% 
  column_to_rownames("gene") %>% 
  as.matrix()
```


## On T cell signatures

```{r}
exh_gs_vec <- c(
  gs_vec[str_subset(names(gs_vec), "TCELL|T_CELL")], 
  exh_gs_vec
)
```

```{r}
ssgsea_exh_df <- run_or_readRDS(
  file.path(file_cache_dir, "ssgsea_exh_df.rds"), 
  function() {
    gsva(
      log1p_cpm_mtx[,str_subset(colnames(log1p_cpm_mtx), "CD8|T cell")],
      gset.idx.list = c(exh_gs_vec, prog_vs_term_exh),
      method = "ssgsea",
      kcdf = "Gaussian",
      min.sz = 3
    ) %>%
      as_tibble(rownames = "gs_name")
  }
)
ssgsea_exh_tidy <- ssgsea_exh_df %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_celltype", values_to = "score") %>% 
  separate(sample_celltype, into = c("sample_ID", "cell_type"), sep = "_") %>% 
  dplyr::rename(NAME = gs_name) %>% 
  group_by(NAME, cell_type) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup() 
```

## On more

### On SPP1+ signature

```{r}
wei_signature_df <- 
  read_excel(here("data", "signatures", "Wei_supptable3.xlsx"),
             sheet = "3-5", 
             skip = 4)
wei_signature_df2 <- 
  read_excel(here("data", "signatures", "Wei_supptable3.xlsx"),
             sheet = "3-3", 
             skip = 2)
```

```{r}
grab_genes_for_sig <- function(df, colstr) {
  df %>% 
    filter(!is.na(get(colstr))) %>% 
    .[[colstr]]
}
wei_gs_vec <- list(
  "SPP1+TAMs" = grab_genes_for_sig(wei_signature_df, "SPP1+TAMs"), 
  "M1" = grab_genes_for_sig(wei_signature_df2, "M1 Macrophage Polarization"), 
  "M2" = grab_genes_for_sig(wei_signature_df2, "M2 Macrophage Polarization"), 
  "Pro-inflamm" = grab_genes_for_sig(wei_signature_df2, "Pro inflammatory"), 
  "Anti-inflammatory" = grab_genes_for_sig(wei_signature_df2, "Anti-inflammatory")
)
```

```{r}
ssgsea_more_df <- run_or_readRDS(
  file.path(file_cache_dir, "ssgsea_more_df.rds"), 
  function() {
    gsva(
      log1p_cpm_mtx,
      gset.idx.list = c(wei_gs_vec),
      method = "ssgsea",
      kcdf = "Gaussian",
      min.sz = 3
    ) %>%
      as_tibble(rownames = "gs_name")
  }
)
ssgsea_more_tidy <- ssgsea_more_df %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_celltype", values_to = "score") %>% 
  separate(sample_celltype, into = c("sample_ID", "cell_type"), sep = "_") %>% 
  dplyr::rename(NAME = gs_name) %>% 
  group_by(NAME, cell_type) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup() 
```

### Plot results

```{r, fig.asp = 1/2, fig.width = 6}
ssgsea_exh_tidy %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  named_group_split(NAME) %>% 
  imap(~{
    ..1 %>% 
      ggplot() +
      aes(mean_hypoxia, zscore) +
      geom_point(size = 2) +
      stat_smooth(method = "lm") + 
      stat_cor(method = "pearson") + 
      facet_grid( ~ cell_type) + 
      labs(title = str_wrap(str_replace_all(..2, "_", " "), width = 20))
  })
```

```{r, fig.asp = 1/1.618, fig.width = 6}
ssgsea_exh_tidy %>% 
  filter(cell_type == "CD8+ T") %>% 
  pivot_wider(id_cols = sample_ID, 
              names_from = NAME, 
              values_from = zscore) %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  mutate(GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13 = 
           GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13_UP - GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13_DN) %>% 
  ggplot() + 
  aes(mean_hypoxia, GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13) + 
  geom_point(size = 2) + 
  stat_smooth(method = "lm") + 
  stat_cor(method = "pearson") + 
  labs(y = "Signature score", 
       title = str_replace_all("GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13", "_", " ") %>% str_wrap(width = 30))

```

```{r}
ssgsea_more_tidy %>% 
  filter(cell_type == "TAM") %>% 
  left_join(mean_hypoxia_df, by = "sample_ID") %>% 
  ggplot() + 
  aes(mean_hypoxia, zscore) + 
  geom_point(size = 2) + 
  stat_smooth(method = "lm") + 
  stat_cor(method = "pearson") + 
  facet_wrap(~NAME, scale = "free")
```

## run GSEA using liger

```{r}
library(liger)
```

```{r}
estimate_list <- pb_ols %>% 
  filter(term == "mean_hypoxia") %>%
  arrange(desc(estimate)) %>%
  named_group_split(cell_type) %>%
  map( ~ set_names(.$estimate, .$gene))
```

```{r}
names(estimate_list)
```


## Hypoxia sigs

```{r}
gsea_hypoxia_df <- run_or_readRDS(
  file.path(file_cache_dir, "gsea_hypoxia_hi.v.lo_hypoxia_gs.rds"), 
  function() {
    gsea_list <- estimate_list %>% 
      map(iterative.bulk.gsea, set.list = c(hypoxia_vec, gs_vec), power = 0)
    gsea_df <- gsea_list %>% 
      map_dfr(as_tibble, rownames = "gs_name", .id = "treatment")
  }
)
```


```{r}
gsea_hypoxia_mtx_df <- gsea_hypoxia_df %>% 
  group_by(treatment, gs_name) %>% 
  pivot_wider(id_cols = gs_name, 
              names_from = treatment, 
              values_from = sscore, 
              values_fill = NA) %>% 
  drop_na() %>% 
  column_to_rownames("gs_name")
gsea_hypoxia_mtx <- gsea_hypoxia_mtx_df %>% 
  as.matrix()

```

```{r, fig.asp = 1/2, fig.width = 42}
ht <- Heatmap(t(gsea_hypoxia_mtx),
              height = unit(1, "in"),
              name = "sscore",
              column_dend_reorder = T, 
              show_column_names = T)
draw(ht)
```


```{r, fig.asp = 1/2, fig.width = 5}
ht <- Heatmap(t(gsea_hypoxia_mtx),
              col = colorRamp2(breaks = c(-3, 0, 4), 
                               colors = c("pink", "white", "black")), 
              height = unit(1.2, "in"),
              name = "sscore",
              column_dend_reorder = T, 
              show_column_names = F)
draw(ht)
```

