---
title: "Sade-feldman analysis - signature scoring"
execute: 
  freeze: auto
  message: false
  warning: false
---

## Load data into R

```{r}
library(Matrix)
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
source(here("lib", "single_cell", "helper_Seurat.R"))

# document specific packages: 
library(Seurat)
library(UCell)
library(Matrix)

file_cache_dir <- file.path(cache_dir, "external-sc-sade-feldman")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

## Seurat workflow

```{r}
so <- run_or_readRDS(
  here("results", "sf_so_pre_processed.rds"), 
  function() {
    so <- readRDS(file.path(data_dir, "scRNAseq", "sf_so_pre_raw.rds"))
    so <- so %>%
      FindVariableFeatures() %>%
      ScaleData() %>%
      RunPCA()
    so <- so %>%
      FindNeighbors() %>%
      RunUMAP(dims = 1:30)
    for (res in c(0.05, 0.1, 0.2, 0.25, 0.3, 0.4, 0.5, 0.6)){
      so <- so %>% FindClusters(resolution = res)
    }
    return(so)
  }
)
```

## Set up some data frames

```{r}
res0.6_clustering <- c(
  "0" = "CD8+ T", 
  "1" = "CD4+ T", 
  "2" = "CD8+ T", 
  "3" = "CD4+ T", 
  "4" = "Treg", 
  "5" = "NK", 
  "6" = "CD8+ T", 
  "7" = "Myeloid", 
  "8" = "B cells", 
  "9" = "CD8+ T", 
  "10" = "B cells", 
  "11" = "B cells", 
  "12" = "Myeloid"
)
```


```{r}
metadata_df <- get_metadata_from_so(so) %>% 
  mutate(my_clustering = res0.6_clustering[as.character(RNA_snn_res.0.6)])

sample_metadata_df <- metadata_df %>% 
  group_by(sample_ID) %>% 
  summarise(ncells = n())
```

```{r}
sample_metadata_df %>% 
  ggplot() + 
  aes(sample_ID, ncells) + 
  geom_col() + 
  rotate_x_text()

metadata_df %>% ggplot() + 
  aes(sample_ID, nFeature_RNA) + 
  geom_violin(scale ="width") + 
  rotate_x_text()
```

```{r}
plot_umap(metadata_df, color_str = "nFeature_RNA", pointsize = 2.1)
plot_umap(metadata_df, color_str = "RNA_snn_res.0.05", pointsize = 2.1)
plot_umap(metadata_df, color_str = "RNA_snn_res.0.1", pointsize = 2.1)
plot_umap(metadata_df, color_str = "immune_type_general", pointsize = 2.1) + 
  theme(legend.position = "bottom")
plot_umap(metadata_df, color_str = "sample_ID", pointsize = 2.1) + 
  theme(legend.position = "none")
```

```{r}
plot_umap(metadata_df, color_str = "response", pointsize = 2.1) 
plot_umap(metadata_df, color_str = "therapy", pointsize = 2.1) 
plot_umap(metadata_df, color_str = "sample_ID", pointsize = 2.1) 
```


```{r}
palette_my_clustering <- c(
  "NK" = "#592941", 
  "CD8+ T" = "darkgreen", 
  "Treg" = "#618985", 
  "CD4+ T" = "#A6D3A0", 
  "Myeloid" = "#B80C09", 
  "B cells" = "#1C77C3"
)

plot_umap(metadata_df, color_str = "my_clustering", pointsize = 2.1) + 
  scale_color_manual(values = palette_my_clustering)
```


```{r}
gene_expr_df <- get_gene_expr_from_so(so,assay = "RNA", gene_vec = c("CD8A", "CD4", "FOXP3", "CD3E", "MS4A1", "CD68"))
metadata_df %>% 
  left_join(gene_expr_df, by = "cell") %>% 
  plot_umap(color_str = "CD8A", pointsize = 2.1)
metadata_df %>% 
  left_join(gene_expr_df, by = "cell") %>% 
  plot_umap(color_str = "CD4", pointsize = 2.1)
metadata_df %>% 
  left_join(gene_expr_df, by = "cell") %>% 
  plot_umap(color_str = "FOXP3", pointsize = 2.1)
metadata_df %>% 
  left_join(gene_expr_df, by = "cell") %>% 
  plot_umap(color_str = "MS4A1", pointsize = 2.1)
metadata_df %>% 
  left_join(gene_expr_df, by = "cell") %>% 
  plot_umap(color_str = "CD68", pointsize = 2.1)
```


# UCell signature scoring

## Hallmarks and Bagaev

```{r}
signature_df <- read_excel(
  file.path(data_dir, "signatures", "1-s2.0-S1535610821002221-mmc2.xlsx"), 
  sheet = "Gene Description", 
  skip = 1
) %>% mutate(Gene = str_replace(Gene, "-", "."))

signature_list <- signature_df %>% 
  named_group_split(`Gene signature`) %>% 
  map(~str_replace(.$Gene, "\\.", "-")) 
```

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
select_gs <- c(
  "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
  "HALLMARK_INTERFERON_ALHPA_RESPONSE", 
  "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", 
  "HALLMARK_ANGIOGENESIS", 
  "HALLMARK_HYPOXIA"
)
select_gs_vec <- gs_vec[select_gs]
```

```{r}
ucell_h <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_stromal_scores.rds"), 
  function() {
    ScoreSignatures_UCell(so@assays$RNA@data,
                          features = c(signature_list, select_gs_vec))
  }
)
```

Warning: The following genes were not found and will be
                        imputed to exp=0:
* CXCL8,VSIR,TRBC1,MARCHF1,WARS1,CCN1,CCN2,P3H1,PRSS2,CAVIN1,CAVIN3,CCN5,ERO1A,LARGE1,NOCT,STBD1

## Benci et al. 

```{r}
ifng_gs <- read_csv(file.path(data_dir, "signatures", "ifng_gs.csv"))$IFNG.GS
isg_rs <- read_csv(file.path(data_dir, "signatures", "isg_rs.csv"))$ISG.RS

benci_list <- list(
  ISG.RS = isg_rs, 
  IFNG.GS = ifng_gs
)
```

```{r}
ucell_benci <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_stromal_benci.rds"), 
  function() {
    ScoreSignatures_UCell(so@assays$RNA@data,
                          features = c(benci_list))
  }
)
```

Warning: The following genes were not found and will be
                        imputed to exp=0:
* PRIC285

# Hypoxia stratification

```{r, fig.asp = 1/1.618, fig.width = 5}
ucell_h %>% 
  as_tibble(rownames = "cell") %>% 
  left_join(metadata_df, by = "cell") %>% 
  ggplot() + 
  aes(sample_ID, HALLMARK_HYPOXIA_UCell, 
      fill = sample_ID) + 
  geom_violin(scale = "width", 
              draw_quantiles = 0.5, 
              color = "white", 
              size = 0.1) + 
  rotate_x_text() + 
  remove_x_spine() + 
  scale_fill_manual(values = palette_compartment) + 
  theme(legend.position = "none")
```

## Sample level

```{r, fig.width = 4, fig.height = 3}
hypoxia_all_df <- ucell_h %>% 
  as_tibble(rownames = "cell") %>% 
  left_join(metadata_df, by = "cell") %>% 
  group_by(sample_ID) %>% 
  summarise(mean_hypoxia = mean(HALLMARK_HYPOXIA_UCell), 
            n = n()) %>% 
  mutate(mean_hypoxia = as.vector(scale(mean_hypoxia)), 
         sd_hypoxia = sd(mean_hypoxia)) %>% 
  mutate(hypoxia_hi.v.lo = case_when(
    mean_hypoxia > quantile(mean_hypoxia, 0.667) ~ "hypoxic", 
    mean_hypoxia > quantile(mean_hypoxia, 0.333) ~ "mid", 
    TRUE ~ "non-hypoxic"
  )) %>% 
  mutate(hypoxia_hi.v.lo = factor(hypoxia_hi.v.lo, 
                                  levels = c("non-hypoxic", "mid", "hypoxic")))
hypoxia_all_df %>% 
  ggplot() + 
  aes(fct_reorder(sample_ID, mean_hypoxia), mean_hypoxia) + 
  annotate("rect", xmin = -Inf, xmax = 6.5, 
           ymin = -Inf, ymax = Inf, 
           fill = palette_hypoxia_state[["low"]], 
           alpha = 0.3) + 
  annotate("rect", xmin = 13.5, xmax = Inf, 
           ymin = -Inf, ymax = Inf, 
           fill = palette_hypoxia_state[["high"]], 
           alpha = 0.3) + 
  geom_point(size = 1) + 
  geom_segment(aes(
    xend = sample_ID,
    y = mean_hypoxia - sd_hypoxia,
    yend = mean_hypoxia + sd_hypoxia
  )) +
  rotate_x_text() +
  remove_x_spine() + 
  labs(x = "Sample", y = "mean hypoxia score") 
```

# Cell types proportions by hypoxia

```{r}
immsubtype_prop_df <- metadata_df %>% 
  group_by(sample_ID, my_clustering) %>% 
  count() %>% 
  group_by(sample_ID) %>% 
  mutate(total = sum(n), proportion = n / total)
```


```{r, fig.asp = 1/3, fig.width = 8}
immsubtype_prop_df %>% 
  left_join(hypoxia_all_df, by = "sample_ID") %>% 
  ggplot() + 
  aes(hypoxia_hi.v.lo, proportion) + 
  geom_boxplot(outlier.size = 0, outlier.stroke = 0) + 
  geom_jitter(size = 3, width = 0.2) + 
  facet_wrap(~my_clustering, ncol = 8, scale = "free_y") +
  rotate_x_text()
```


```{r, fig.asp = 1/3, fig.width = 8}
immsubtype_prop_df %>% 
  left_join(hypoxia_all_df, by = "sample_ID") %>% 
  mutate(hypoxia_hi.v.lo = 
           case_when(
             hypoxia_hi.v.lo == "non-hypoxic" ~ "low", 
             hypoxia_hi.v.lo == "mid" ~ "mid", 
             hypoxia_hi.v.lo == "hypoxic" ~ "high"
           )) %>% 
  mutate(hypoxia_hi.v.lo = factor(hypoxia_hi.v.lo, levels = c(
    "low", "mid", "high"
  ))) %>% 
  ggplot() + 
  aes(hypoxia_hi.v.lo, proportion) + 
  geom_col(aes(y = mean, fill = hypoxia_hi.v.lo), 
           data = ~..1 %>% 
             group_by(my_clustering, hypoxia_hi.v.lo) %>% 
             summarise(mean = mean(proportion))) + 
  geom_jitter(size = 3, width = 0.2) + 
  facet_wrap(~my_clustering, ncol = 8, scale = "free_y") +
  rotate_x_text() + 
  remove_x_spine() + 
  scale_fill_manual(values = palette_hypoxia_state) + 
  stat_compare_means(comparisons = list(c("low", "high")))+ 
  scale_y_continuous(limits = ~c(0, ..1[2]*1.2), 
                     expand = c(0, 0)) + 
  labs(y = "Frequency")
```


# Characterizing Hypoxia 

## Expression of hypoxia-related transcription factors

```{r, fig.asp = 1/4, fig.width = 10}
hypoxia_genes <- c("EPAS1", "HIF1A", "VEGFA", "FLT1", "KDR")
hypoxia_gene_df <- get_gene_expr_from_so(so, gene_vec = hypoxia_genes)

metadata_df %>% 
  left_join(hypoxia_gene_df, by = "cell") %>%
  pivot_longer(cols = hypoxia_genes) %>% 
  group_by(sample_ID, my_clustering, name) %>% 
  summarise(mean_value = mean(value)) %>% 
  drop_na(mean_value) %>% 
  left_join(hypoxia_all_df, by = "sample_ID") %>% 
  ggplot() + 
  aes(my_clustering, mean_value) + 
  geom_col(aes(y = mean_mean_value, fill = my_clustering), 
           data = ~..1 %>% 
             group_by(my_clustering, name) %>% 
             summarise(mean_mean_value = mean(mean_value))) +
  geom_jitter(width = 0.2, size = 1) + 
  scale_fill_manual(values = palette_my_clustering) +
  facet_wrap( ~ name, ncol = 5, scale = "free_x") +
  theme(legend.position = "none") + 
  coord_flip() 
```

## Hypoxia and response

```{r}
metadata_df %>% 
  group_by(sample_ID) %>% 
  summarise(response = unique(response), 
            therapy = unique(therapy)) %>% 
  left_join(hypoxia_all_df, by = "sample_ID") %>% 
  ggplot() + 
  aes(response, mean_hypoxia) + 
  geom_boxplot() + 
  geom_jitter(width = 0.2, size = 3, 
              aes(shape = therapy, color = hypoxia_hi.v.lo)) + 
  scale_color_manual(values = unname(palette_hypoxia_state))
```

## Saving to so

```{r}
if ("hypoxia_hi.v.lo" %in% colnames(metadata_df))
  metadata_df <- metadata_df %>% dplyr::select(-hypoxia_hi.v.lo)

metadata_to_add <- metadata_df %>% 
  left_join(hypoxia_all_df, by = "sample_ID") %>% 
  dplyr::select(cell, my_clustering, hypoxia_hi.v.lo) %>% 
  column_to_rownames("cell")
so <- AddMetaData(so, metadata = metadata_to_add)
```

```{r, eval = F}
saveRDS(so, here("results", "sf_so_pre_processed.rds"))
```

# pb

```{r}
make_specific_pb_hypoxia <- function(so, cluster_str, ident_str) {
  so_subset <- subset(so %>% set_idents(cluster_str),
                      ident = ident_str)
  so_subset@meta.data$sample_ID <-
    factor(so_subset@meta.data$sample_ID)
  pb <- make_pseudobulk(
    so_subset,
    split_var = NULL,
    sample_var = "sample_ID",
    vars = c("hypoxia_hi.v.lo")
  )
  return(pb)
}

pb_list <- run_or_readRDS(
  file.path(file_cache_dir, "pb.rds"), 
  function() {
    pb_imm_list <-
      so@meta.data$my_clustering %>% unique() %>%
      set_names() %>%
      map( ~ make_specific_pb_hypoxia(so, "my_clustering", ..1))
  }
)
```

# Setting up modeling

## Attempt: y=log1p(CPM), OLS

```{r}
pb_df <- pb_list %>% 
  map_dfr(as_tibble, rownames = "sample_ID", .id = "my_clustering") %>% 
  separate(sample_ID, sep = "_", into = c("state", "patient_ID", "hypoxia_hi.v.lo")) %>% 
  unite(sample_ID, state, patient_ID, remove = F)

pb_tidy <- pb_df %>% 
  pivot_longer(cols = 6:ncol(.), 
               names_to = "gene", 
               values_to = "count") %>% 
  mutate(count = ifelse(is.na(count), 0, count)) %>% 
  group_by(sample_ID, state, my_clustering) %>% 
  mutate(total_TPM = sum(count), reTPM = count / total_TPM * 1e6) %>% 
  ungroup() %>% 
  mutate(log1p_reTPM = log10(reTPM + 1)) %>% 
  left_join(select(hypoxia_all_df, sample_ID, mean_hypoxia), 
            by = c("sample_ID"))
```

```{r, eval = FALSE}
saveRDS(pb_tidy, here("results", "sade-feldman_sc_pb_tidy.rds"))
```

### Model

For each cell type: 

```
log1p_CPM ~ hypoxia
```

And we'll only consider genes with abs(log2FC) > 0.1. 

```{r}
test_genes_in_cell_type <- function(my_pb_tidy, thresh = 0.1) {
  log2fc_df <- my_pb_tidy %>% 
    group_by(gene, hypoxia_hi.v.lo) %>% 
    summarise(mean_log1p_reTPM = mean(log1p_reTPM)) %>% 
    pivot_wider(id_cols = gene, 
                names_from = hypoxia_hi.v.lo, 
                values_from = mean_log1p_reTPM) %>% 
    mutate(log2FC_high_over_low = log2(`hypoxic` / `non-hypoxic`))
  
  select_genes <- log2fc_df %>% 
    filter(abs(log2FC_high_over_low) > thresh) %>% 
    .$gene
  
  my_pb_tidy %>% 
    filter(gene %in% select_genes) %>% 
    group_by(gene) %>% 
    summarise(lm_out = broom::tidy(lm(log1p_reTPM ~ mean_hypoxia))) %>% 
    unnest(lm_out) %>% 
    left_join(log2fc_df, by = "gene")
}

```

```{r}
pb_tidy_split <- pb_tidy %>% 
  named_group_split(my_clustering)

model_out <- run_or_readRDS(
  file.path(file_cache_dir, "model_out.rds"), 
  function() {
    pb_tidy_split %>% 
      map_dfr(test_genes_in_cell_type, .id = "my_clustering")
  }
)
```

### Save

```{r, eval = FALSE}
saveRDS(model_out, here("results", "sade-feldman-sc-analysis-hypoxia-pb-gene-expr-modeling-simplified.rds"))
```

# pb ols results 

```{r}
pb_ols <- model_out
pb_cpm_wide <- pb_tidy %>% 
  pivot_wider(id_cols = c(my_clustering, sample_ID), 
              names_from = gene, 
              values_from = log1p_reTPM)
mean_hypoxia_df <- hypoxia_all_df
```


```{r}
plot_pb_ols_ranks <- function(my_pb_ols) {
  my_pb_ols %>% 
    arrange(estimate) %>%
    mutate(rank = row_number()) %>%
    ggplot() +
    aes(rank, estimate) +
    geom_hline(yintercept = 0, linetype = "dotted") +
    geom_line() +
    geom_segment(
      aes(
        x = rank,
        xend = rank,
        y = 0,
        yend = estimate,
        color = category
      ),
      data = ~ ..1 %>% drop_na(category)
    ) +
    geom_point(aes(color = category, shape = p.value < 0.05),
               data = ~ ..1 %>% drop_na(category),
               size = 3) +
    geom_text_repel(
      aes(label = gene, color = category),
      data = ~ ..1 %>% drop_na(category),
      force = 5,
      hjust = 1
    )  + 
    scale_shape_manual(values = c("TRUE" = 19, "FALSE" = 4)) + 
    theme(aspect.ratio = 1)
}
```

### CD8+ T cells

```{r}
exhaustion_genes <- c("ENTPD1", "CXCL13", "CD101", "HAVCR2", 
                      "LAG3", "PDCD1")#, "CTLA4", "TIGIT")
activation_genes <- c("NFAT5", "TNFSF9")#, "KLRG1", "STAT1", "TAP1")
homing_genes <- c("CCR7")
hypoxia_genes <- c("HIF1A", "EPAS1")

pb_ols %>% 
  ungroup() %>% 
  filter(my_clustering == "CD8+ T") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% exhaustion_genes ~ "exhaustion", 
    gene %in% activation_genes ~ "activation", 
    # gene %in% homing_genes ~ "homing", 
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    TRUE ~ NA_character_
  )) %>% 
  plot_pb_ols_ranks() + 
  labs(title = "Sade-Feldman: CD8+")
```

### Tregs

```{r}
exhaustion_genes <- c("ENTPD1", "CXCL13", "CD101", "HAVCR2", 
                      "LAG3", "PDCD1")#, "CTLA4", "TIGIT")
activation_genes <- c("NFAT5", "TNFSF9")#, "KLRG1", "STAT1", "TAP1")
homing_genes <- c("CCR7")
hypoxia_genes <- c("HIF1A", "EPAS1")

pb_ols %>% 
  ungroup() %>% 
  filter(my_clustering == "Treg") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% exhaustion_genes ~ "exhaustion", 
    gene %in% activation_genes ~ "activation", 
    # gene %in% homing_genes ~ "homing", 
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    TRUE ~ NA_character_
  )) %>% 
  plot_pb_ols_ranks() + 
  labs(title = "Sade-Feldman: Tregs")
```

```{r, fig.asp = 1/4, fig.width = 10}
genes_to_show <- c("ENTPD1", "LAG3", "CXCL13", "HAVCR2")
exh_cd8_df <- pb_cpm_wide %>% 
  filter(my_clustering == "Treg") %>% 
  select(sample_ID, genes_to_show) %>% 
  left_join(mean_hypoxia_df)
exh_cd8_df %>% 
  pivot_longer(cols = genes_to_show) %>% 
  ggplot() + 
  aes(mean_hypoxia, value) + 
  geom_point(size = 2) + 
  stat_smooth(method = "lm", color = "black", se = F) + 
  facet_wrap(~name, scale = "free", ncol = 4) + 
  # theme(legend.position = "top",
  #       legend.direction = "vertical") + 
  scale_color_manual(values = palette_immune_state) + 
  stat_cor(method = "spearman") + 
  labs(y = "normalized expr")
```

### Myeloid

```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
immunosuppressive_genes <- c("CD163", "SPP1", "CCL2", "CSF1", "CSF1R", "ARG1")
proinflammatory_genes <- c("IL6", "C1QB", "NOS2", "TLR2", "TLR4", "TNF", "IL1B")
adhesion_genes <- c("VCAM1")
cytokines_and_chemokines <- c("IL33", "CXCL6", 
                              # "CXCL10", "CXCL11",
                              "CXCL8", "CXCL2")

pb_ols %>% 
  ungroup() %>% 
  filter(my_clustering == "Myeloid") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    # gene %in% hypoxia_genes ~ "hypoxia response" ,
    gene %in% immunosuppressive_genes ~ "immunosuppressive", 
    gene %in% proinflammatory_genes ~ "proinflammatory", 
    # gene %in% adhesion_genes ~ "adhesion", 
    gene %in% cytokines_and_chemokines ~ "cytokines and chemokines", 
    TRUE ~ NA_character_
  )) %>% 
  plot_pb_ols_ranks() + 
  labs(title = "Sade-Feldman: TAMs")
```

```{r, fig.asp = 1/4, fig.width = 10}
genes_to_show <- c("IL1B", "CXCL2", "SPP1")
gene_expr_df <- pb_cpm_wide %>% 
  filter(my_clustering == "Myeloid") %>% 
  select(sample_ID, genes_to_show) %>% 
  left_join(mean_hypoxia_df)
gene_expr_df %>% 
  pivot_longer(cols = genes_to_show) %>% 
  ggplot() + 
  aes(mean_hypoxia, value) + 
  geom_point(size = 2) + 
  stat_smooth(method = "lm", color = "black", se = F) + 
  facet_wrap(~name, scale = "free", ncol = 4) + 
  # theme(legend.position = "top",
  #       legend.direction = "vertical") + 
  scale_color_manual(values = palette_immune_state) + 
  stat_cor(method = "spearman") + 
  labs(y = "normalized expr")
```

# Gene signatures

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
hypoxia_vec <- readRDS(file.path(results_dir, "hypoxia_gs_vec.rds"))
prog_vs_term_exh <- readRDS(here("results", "gs_vec_prog.vs.term_exh.rds"))
exh_gs_vec <- readRDS(file.path(results_dir, "exh_gs_vec.rds"))
```

# Run ssGSEA

```{r}
library(GSVA)
```

```{r}
log1p_tpm_mtx <- pb_tidy %>% 
  pivot_wider(id_cols = gene, 
              names_from = c(sample_ID, my_clustering), 
              values_from = log1p_reTPM, 
              values_fill = 0) %>% 
  column_to_rownames("gene") %>% 
  as.matrix()
```


## On T cell signatures

```{r}
exh_gs_vec <- c(
  gs_vec[str_subset(names(gs_vec), "TCELL|T_CELL")], 
  exh_gs_vec
)
```

```{r}
ssgsea_exh_df <- run_or_readRDS(
  file.path(file_cache_dir, "ssgsea_exh_df.rds"), 
  function() {
    gsva(
      log1p_tpm_mtx[,str_subset(colnames(log1p_tpm_mtx), "CD8|T cell")],
      gset.idx.list = c(exh_gs_vec, prog_vs_term_exh),
      method = "ssgsea",
      kcdf = "Gaussian",
      min.sz = 3
    ) %>%
      as_tibble(rownames = "gs_name")
  }
)
ssgsea_exh_tidy <- ssgsea_exh_df %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_celltype", values_to = "score") %>% 
  separate(sample_celltype, into = c("state", "patient_ID", "cell_type"), sep = "_") %>% 
  unite(sample_ID, state, patient_ID, remove = F) %>% 
  dplyr::rename(NAME = gs_name) %>% 
  group_by(NAME, cell_type) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup() 
```

## On more

### On SPP1+ signature

```{r}
wei_signature_df <- 
  read_excel(here("data", "signatures", "Wei_supptable3.xlsx"),
             sheet = "3-5", 
             skip = 4)
wei_signature_df2 <- 
  read_excel(here("data", "signatures", "Wei_supptable3.xlsx"),
             sheet = "3-3", 
             skip = 2)
```

```{r}
grab_genes_for_sig <- function(df, colstr) {
  df %>% 
    filter(!is.na(get(colstr))) %>% 
    .[[colstr]]
}
wei_gs_vec <- list(
  "SPP1+TAMs" = grab_genes_for_sig(wei_signature_df, "SPP1+TAMs"), 
  "M1" = grab_genes_for_sig(wei_signature_df2, "M1 Macrophage Polarization"), 
  "M2" = grab_genes_for_sig(wei_signature_df2, "M2 Macrophage Polarization"), 
  "Pro-inflamm" = grab_genes_for_sig(wei_signature_df2, "Pro inflammatory"), 
  "Anti-inflammatory" = grab_genes_for_sig(wei_signature_df2, "Anti-inflammatory")
)
```

```{r}
ssgsea_more_df <- run_or_readRDS(
  file.path(file_cache_dir, "ssgsea_more_df.rds"), 
  function() {
    gsva(
      log1p_tpm_mtx,
      gset.idx.list = c(wei_gs_vec),
      method = "ssgsea",
      kcdf = "Gaussian",
      min.sz = 3
    ) %>%
      as_tibble(rownames = "gs_name")
  }
)
ssgsea_more_tidy <- ssgsea_more_df %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_celltype", values_to = "score") %>% 
  separate(sample_celltype, into = c("state", "patient_ID", "cell_type"), sep = "_") %>% 
  unite(sample_ID, state, patient_ID, remove = F) %>% 
  dplyr::rename(NAME = gs_name) %>% 
  group_by(NAME, cell_type) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup() 
```

## Plot results

### CD8+ T

```{r, fig.asp = 1, fig.width = 4}
ssgsea_exh_tidy %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  named_group_split(NAME) %>% 
  imap(~{
    ..1 %>% 
      ggplot() +
      aes(mean_hypoxia, zscore) +
      geom_point(size = 2) +
      stat_smooth(method = "lm") + 
      stat_cor(method = "pearson") + 
      facet_grid( ~ cell_type) + 
      labs(title = str_wrap(str_replace_all(..2, "_", " "), width = 20))
  })
```

```{r, fig.asp = 1, fig.width = 4}
ssgsea_exh_tidy %>% 
  filter(cell_type == "CD8+ T") %>% 
  pivot_wider(id_cols = sample_ID, 
              names_from = NAME, 
              values_from = zscore) %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  mutate(GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13 = 
           GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13_UP - GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13_DN) %>% 
  ggplot() + 
  aes(mean_hypoxia, GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13) + 
  geom_point(size = 2) + 
  stat_smooth(method = "lm") + 
  stat_cor(method = "pearson") + 
  labs(y = "Signature score", 
       title = str_replace_all("GSE41867_DAY15_EFFECTOR_VS_DAY30_EXHAUSTED_CD8_TCELL_LCMV_CLONE13", "_", " ") %>% str_wrap(width = 30))

```

### Myeloid

```{r, fig.asp = 1 / 1.618, fig.width = 7}
ssgsea_more_tidy %>% 
  filter(cell_type == "Myeloid") %>% 
  left_join(mean_hypoxia_df, by = "sample_ID") %>% 
  ggplot() + 
  aes(mean_hypoxia, zscore) + 
  geom_point(size = 2) + 
  stat_smooth(method = "lm") + 
  facet_wrap(~NAME, scale = "free")
```

## run GSEA using liger

```{r}
library(liger)
```

```{r}
estimate_list <- pb_ols %>% 
  filter(term == "mean_hypoxia") %>%
  arrange(desc(estimate)) %>%
  named_group_split(my_clustering) %>%
  map( ~ set_names(.$estimate, .$gene))
```

```{r}
names(estimate_list)
```


### Hypoxia sigs

```{r}
gsea_hypoxia_df <- run_or_readRDS(
  file.path(file_cache_dir, "gsea_hypoxia_hi.v.lo_hypoxia_gs.rds"), 
  function() {
    gsea_list <- estimate_list %>% 
      map(iterative.bulk.gsea, set.list = c(hypoxia_vec, gs_vec), power = 0)
    gsea_df <- gsea_list %>% 
      map_dfr(as_tibble, rownames = "gs_name", .id = "treatment")
  }
)
```


```{r}
gsea_hypoxia_mtx_df <- gsea_hypoxia_df %>% 
  group_by(treatment, gs_name) %>% 
  pivot_wider(id_cols = gs_name, 
              names_from = treatment, 
              values_from = sscore, 
              values_fill = NA) %>% 
  drop_na() %>% 
  column_to_rownames("gs_name")
gsea_hypoxia_mtx <- gsea_hypoxia_mtx_df %>% 
  as.matrix()

```

```{r, fig.asp = 1/2, fig.width = 42}
ht <- Heatmap(t(gsea_hypoxia_mtx),
              height = unit(1, "in"),
              name = "sscore",
              column_dend_reorder = T, 
              show_column_names = T)
draw(ht)
```


```{r, fig.asp = 1/2, fig.width = 5}
ht <- Heatmap(t(gsea_hypoxia_mtx),
              col = colorRamp2(breaks = c(-3, 0, 4), 
                               colors = c("pink", "white", "black")), 
              height = unit(1.2, "in"),
              name = "sscore",
              column_dend_reorder = T, 
              show_column_names = F)
draw(ht)
```

