---
title: "DGE hypoxia for each cell type, simplified THIS IS THE ONE"
execute: 
  message: false
  warning: false
  freeze: auto
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
source(here("lib", "single_cell", "helper_Seurat.R"))

library(Seurat)
library(Matrix)
library(Matrix.utils)
library(liger)
library(edgeR)

file_cache_dir <- file.path(cache_dir, "sc-analysis-hypoxia-pb-gene-expr-modeling-simplified")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

# Read in data

## data frames

```{r}
immune_score_df <- readRDS(file.path(results_dir, "sc_immune_prop_df.rds"))
aucell_df <- readRDS(file.path(data_dir, "scRNAseq", "adata_aucell_scores_df.rds"))
mean_hypoxia_df <- readRDS(file.path(data_dir, "scRNAseq", "mean_hypoxia_df.rds"))
obs_df <- read_csv(file.path(data_dir, "scRNAseq", "adata_pre_obs_2024-04-07.csv")) %>% 
  left_join(aucell_df, by = c("cell_label" = "cell")) %>% 
  left_join(mean_hypoxia_df)
immune_prop_df <- read_tsv(file.path(results_dir, "sc_immune_prop_df.tsv")) %>% 
  mutate(immune_hi.v.lo = factor(immune_hi.v.lo, levels = c(
    "immune low", "immune mid", "immune high"
  )))
```

## Seurat objects 

Filter Seurat objects to only include cells with nFeature_RNA > 1000. 

```{r}
read_in_so_tumor <- function() {
  so_tumor <- readRDS(file.path(cache_dir, "sc-processing-sctransform-tumor", "so_tumor.rds"))
  so_tumor <- AddMetaData(
    so_tumor,
    metadata = so_tumor@meta.data %>%
      as_tibble(rownames = "cell") %>%
      left_join(mean_hypoxia_df, by = "sample_ID") %>%
      dplyr::select(cell, hypoxia_hi.v.lo, mean_hypoxia) %>%
      column_to_rownames("cell")
  )
  return(so_tumor)
}

read_in_so_stromal <- function() {
  so_stromal <- readRDS(file.path(cache_dir, "sc-processing-sctransform-stromal", "so_stromal.rds")) %>% 
    subset(subset = nFeature_RNA > 1000)
  so_stromal <- AddMetaData(so_stromal, metadata = so_stromal@meta.data %>%
                              as_tibble(rownames = "cell") %>%
                              left_join(mean_hypoxia_df, by = "sample_ID") %>%
                              dplyr::select(cell, hypoxia_hi.v.lo, mean_hypoxia) %>% 
                              column_to_rownames("cell"))
  return(so_stromal)
  
}

read_in_so_imm <- function() {
  so_imm <- readRDS(file.path(cache_dir, "sc-processing-sctransform-immune", "so_immune.rds")) %>% 
    subset(subset = nFeature_RNA > 1000)
  so_imm <- AddMetaData(
    so_imm,
    metadata = so_imm@meta.data %>%
      as_tibble(rownames = "cell") %>%
      left_join(mean_hypoxia_df, by = "sample_ID") %>%
      dplyr::select(cell, hypoxia_hi.v.lo, mean_hypoxia) %>%
      column_to_rownames("cell"
      )
  )
  return(so_imm)
}


```

# Make pseudobulk

```{r}
my_seurat_to_sce <- function (so, split_var = NULL, sample_var, vars) 
{
    seurat_counts <- so@assays$SCT@counts # need to use SCT counts
    seurat_metadata <- so@meta.data %>% dplyr::select(sample_var, 
        vars)
    if (is.null(split_var)) {
        sce_obj <- SingleCellExperiment::SingleCellExperiment(assays = list(counts = seurat_counts), 
            colData = seurat_metadata)
        sce_obj <- qc_sce(sce_obj, sample_var)
        return(sce_obj)
    }
    else {
        seurat_metadata <- so@meta.data %>% dplyr::select(vars, 
            sample_var, split_var)
        sce <- SingleCellExperiment::SingleCellExperiment(assays = list(counts = seurat_counts), 
            colData = seurat_metadata)
        sce_list <- unique(seurat_metadata[[split_var]]) %>% 
            purrr::map(~base::subset(sce, , get(split_var) == 
                ..1)) %>% set_names(unique(seurat_metadata[[split_var]]))
        sce_list <- purrr::map(sce_list, qc_sce, sample_var)
        return(sce_list)
    }
}

my_make_pseudobulk <- function (so, split_var = NULL, sample_var, vars)
{
  sce <- my_seurat_to_sce(so, split_var, sample_var, vars)
  if (is.null(split_var)) {
    groups <- SummarizedExperiment::colData(sce)[, c(sample_var,
                                                     vars)]
    pb <-
      Matrix.utils::aggregate.Matrix(
        Matrix::t(SingleCellExperiment::counts(sce)),
        groupings = groups,
        fun = "sum"
      )
    return(pb)
  }
  else {
    groups <- map(sce, ~ SummarizedExperiment::colData(..1)[,
                                                            c(sample_var, vars)])
    pb_list <-
      map2(
        sce,
        groups,
        ~ Matrix.utils::aggregate.Matrix(
          Matrix::t(SingleCellExperiment::counts(..1)),
          groupings = ..2,
          fun = "sum"
        )
      ) %>% set_names(names(sce))
    return(pb_list)
  }
}
```


```{r}
make_specific_pb_hypoxia <- function(so, cluster_str, ident_str) {
  so_subset <- subset(so %>% set_idents(cluster_str),
                      ident = ident_str)
  so_subset@meta.data$sample_ID <-
    factor(so_subset@meta.data$sample_ID)
  pb <- my_make_pseudobulk(
    so_subset,
    split_var = NULL,
    sample_var = "sample_ID",
    vars = c("hypoxia_hi.v.lo")
  )
  return(pb)
}

pb_imm_list <- run_or_readRDS(
  file.path(cache_dir, 
            "sc-analysis-hypoxia-pb-gene-expr-modeling-simplified",
            "pb_imm_list.rds"), 
  function() {
    so_imm <- read_in_so_imm()
    pb_imm_list <-
      so_imm@meta.data$specific_cluster_immune %>% unique() %>%
      .[-3] %>% # dropping NA
      set_names() %>%
      map( ~ make_specific_pb_hypoxia(so_imm, "specific_cluster_immune", ..1))
  }
)
```

```{r}
pb_imm_list[["Neutrophil"]] <- NULL
pb_imm_list[["Doublets"]] <- NULL
pb_imm_list[["Contaminating"]] <- NULL
pb_imm_list[["Stressed"]] <- NULL
```

```{r}
pb_stromal_list <- run_or_readRDS(
  file.path(cache_dir, 
            "sc-analysis-hypoxia-pb-gene-expr-modeling-simplified",
            "pb_stromal_list.rds"), 
  function() {
    so_stromal <- read_in_so_stromal()
    pb_stromal_list <-
      so_stromal@meta.data$general_cluster_stromal %>% 
      unique() %>%
      set_names() %>%
      map( ~ make_specific_pb_hypoxia(so_stromal, "general_cluster_stromal", ..1))
  }
)
```

```{r}
pb_tumor <- run_or_readRDS(
  file.path(cache_dir, 
            "sc-analysis-hypoxia-pb-gene-expr-modeling-simplified",
            "pb_tumor.rds"), 
  function( ){
    so_tumor <- read_in_so_tumor()
    pb_tumor <- make_pseudobulk(so_tumor,
                                split_var = NULL,
                                sample_var = "sample_ID",
                                vars =  c("hypoxia_hi.v.lo"))
  }
)
```

# Setting up modeling

## Attempt: y=log1p(CPM), OLS

### Prepare data and run analysis

```{r}
select_hypoxia_df <- mean_hypoxia_df %>% 
  dplyr::select(sample_ID, mean_hypoxia)
select_immune_df <- immune_prop_df %>% 
  mutate(immune_prop = prop) %>% 
  dplyr::select(sample_ID, immune_prop)

pb_df <- c(pb_imm_list, pb_stromal_list, list(Tumor = pb_tumor)) %>% 
  map_dfr(as_tibble, rownames = "sample_ID", .id = "cell_type") %>% 
  separate(sample_ID, sep = "_", into = c("sample_ID", "hypoxia_hi.v.lo"))

pb_tidy <- pb_df %>% 
  pivot_longer(cols = 4:ncol(.), 
               names_to = "gene", 
               values_to = "count") %>% 
  mutate(count = ifelse(is.na(count), 0, count)) %>% 
  group_by(sample_ID, cell_type) %>% 
  mutate(total_counts = sum(count), CPM = count / total_counts * 1e6) %>% 
  ungroup() %>% 
  mutate(log1p_CPM = log10(CPM + 1))

pb_tidy <- left_join(pb_tidy, select_hypoxia_df, by = "sample_ID") %>% 
  left_join(select_immune_df, by = "sample_ID") %>% 
  mutate(hypoxia_hi.v.lo = factor(hypoxia_hi.v.lo, levels = c(paste0(
    c("low", "mid", "high"), " hypoxia"
  )), ordered = TRUE))
```

```{r, eval = FALSE}
saveRDS(pb_tidy, here("results", "sc_pb_tidy.rds"))
```


### Model

For each cell type: 

```
log1p_CPM ~ hypoxia
```

And we'll only consider genes with abs(log2FC) > 0.1. 

```{r}
test_genes_in_cell_type <- function(my_pb_tidy, thresh = 0.1) {
  log2fc_df <- my_pb_tidy %>% 
    group_by(gene, hypoxia_hi.v.lo) %>% 
    summarise(mean_log1p_CPM = mean(log1p_CPM)) %>% 
    pivot_wider(id_cols = gene, 
                names_from = hypoxia_hi.v.lo, 
                values_from = mean_log1p_CPM) %>% 
    mutate(log2FC_high_over_low = log2(`high hypoxia` / `low hypoxia`))
  
  select_genes <- log2fc_df %>% 
    filter(abs(log2FC_high_over_low) > thresh) %>% 
    .$gene
  
  my_pb_tidy %>% 
    filter(gene %in% select_genes) %>% 
    group_by(gene) %>% 
    summarise(lm_out = broom::tidy(lm(log1p_CPM ~ mean_hypoxia))) %>% 
    unnest(lm_out) %>% 
    left_join(log2fc_df, by = "gene")
}

```

```{r}
pb_tidy_split <- pb_tidy %>% 
  named_group_split(cell_type)

model_out <- run_or_readRDS(
  file.path(file_cache_dir, "model_out.rds"), 
  function() {
    pb_tidy_split[c(
        "CD8+ T",
        "DC",
        "TAM",
        "Endothelial cells",
        "Fibroblasts",
        "Tumor",
        "NK",
        "Tregs", 
        "Naive-like T"
      )] %>%
      map_dfr(test_genes_in_cell_type, .id = "cell_type")
  }
)
```

## Save

```{r, eval = FALSE}
saveRDS(model_out, here("results", "sc-analysis-hypoxia-pb-gene-expr-modeling-simplified.rds"))
model_out <- readRDS(here("results", 
                  "sc-analysis-hypoxia-pb-gene-expr-modeling-simplified.rds"))
```

## Results

```{r, eval = F}
so_imm <- read_in_so_imm()
```


### Properties

```{r}
model_out %>% 
  group_by(cell_type) %>% 
  count()
```

### CD8+ T

```{r}
model_out %>% 
  filter(cell_type == "CD8+ T") %>% 
  filter(term == "mean_hypoxia") %>% 
  filter(p.value < 0.05) %>% 
  arrange(desc(estimate)) %>% 
  head(20)
```


```{r, fig.asp = 1, eval = F}
metadata_df <- get_metadata_from_so(so_imm)
genes <- c("CD8A", "CD3D", "CD4", "FOXP3", "IGKC", "FCMR", "TRDC", "FCGR3A", 
           "HAVCR2", "CXCL13", "TOX")
gene_expr_df <- get_gene_expr_from_so(so_imm, genes, assay = "SCT")
sample_ID_hypoxia_ordering <- mean_hypoxia_df %>% 
  arrange(mean_hypoxia) %>% 
  .$sample_ID
metadata_df %>% 
  filter(specific_cluster_immune == "CD8+ T") %>% 
  left_join(gene_expr_df, by = "cell") %>% 
  mutate(sample_ID = factor(sample_ID, levels = sample_ID_hypoxia_ordering)) %>% 
  pivot_longer(cols = genes) %>% 
  ggplot() + 
  aes(sample_ID, value) + 
  geom_violin(scale = "width") + 
  rotate_x_text() + 
  facet_grid(name~.) + 
  theme(strip.text.y = element_text(angle = 0))
```

### Tregs

```{r}
model_out %>% 
  filter(cell_type == "Tregs") %>% 
  filter(term == "mean_hypoxia") %>% 
  filter(p.value < 0.05) %>% 
  arrange(desc(estimate)) %>% 
  head(20)
```


```{r, fig.asp = 1, eval = F}
metadata_df <- get_metadata_from_so(so_imm)
genes <- c("CD8A", "CD3D", "CD4", "FOXP3", "IGKC", "FCMR", "TRDC", "FCGR3A")
gene_expr_df <- get_gene_expr_from_so(so_imm, genes, assay = "SCT")
sample_ID_hypoxia_ordering <- mean_hypoxia_df %>% 
  arrange(mean_hypoxia) %>% 
  .$sample_ID
metadata_df %>% 
  filter(specific_cluster_immune == "Tregs") %>% 
  left_join(gene_expr_df, by = "cell") %>% 
  mutate(sample_ID = factor(sample_ID, levels = sample_ID_hypoxia_ordering)) %>% 
  pivot_longer(cols = genes) %>% 
  ggplot() + 
  aes(sample_ID, value) + 
  geom_violin(scale = "width") + 
  rotate_x_text() + 
  facet_grid(name~.)
```

### TAM

```{r}
model_out %>% 
  ungroup() %>% 
  filter(cell_type == "TAM") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  head(20)
```


```{r, fig.asp = 1, eval = F}
metadata_df <- get_metadata_from_so(so_imm)
genes <- c("TRAC", "IGKC", "HBB", "LY6E", "CD3E", "CD14", "CD68", "AIF1", "SPP1", "CD2")
gene_expr_df <- get_gene_expr_from_so(so_imm, assay="SCT", genes)
sample_ID_hypoxia_ordering <- mean_hypoxia_df %>% 
  arrange(mean_hypoxia) %>% 
  .$sample_ID
metadata_df %>% 
  filter(specific_cluster_immune == "TAM") %>% 
  left_join(gene_expr_df, by = "cell") %>% 
  mutate(sample_ID = factor(sample_ID, levels = sample_ID_hypoxia_ordering)) %>% 
  pivot_longer(cols = genes) %>% 
  ggplot() + 
  aes(sample_ID, value) + 
  geom_violin(scale = "width") + 
  rotate_x_text() + 
  facet_grid(name~.) + 
  labs(x = "sample ID, ordered left to right by increasing hypoxia", 
       title = "TAM")
```

### Line plots

```{r}
coinh_receptors <- c("PDCD1", "CTLA4", "TIGIT")
activation_genes <- c("NFAT5")
homing_genes <- c("CCR7")
hypoxia_genes <- c("HIF1A", "EPAS1")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "Tregs") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% coinh_receptors ~ "coinh. receptors", 
    gene %in% activation_genes ~ "activation", 
    gene %in% homing_genes ~ "homing", 
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            hjust = 1) + 
  labs(title = "Tregs", y = "hypoxia effect estimate")
```

```{r}
activation_genes <- c("NFAT5", "CD40LG", 
                      # "TNFRSF9", 
                      "TNFSF9", "KLRG1", "TAP1", "STAT1")

homing_genes <- c("CCR7")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "Naive-like T") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% activation_genes ~ "activation", 
    gene %in% homing_genes ~ "homing", 
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            hjust = 1) + 
  labs(title = "Naive like T", y = "hypoxia effect estimate")
```

```{r}
effector_genes <- c("IFNGR1", "FOSL2", "TNF", "KLF2")
terminal_maturation <- c("CX3CR1")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "NK") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% effector_genes ~ "effector?", 
    gene %in% terminal_maturation ~ "terminal maturation", 
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            hjust = 1) + 
  labs(title = "NK", y = "hypoxia effect estimate")
```

```{r}
exhaustion_genes <- c("ENTPD1", "CXCL13", "CD101", "HAVCR2", 
                      "LAG3", "PDCD1")#, "CTLA4", "TIGIT")
activation_genes <- c("NFAT5", "TNFSF9")#, "KLRG1", "STAT1", "TAP1")
homing_genes <- c("CCR7")
hypoxia_genes <- c("HIF1A", "EPAS1")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "CD8+ T") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% exhaustion_genes ~ "exhaustion", 
    gene %in% activation_genes ~ "activation", 
    gene %in% homing_genes ~ "homing", 
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            force = 5, 
            hjust = 1) + 
  labs(title = "CD8+ T cells")
```

```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
cdc1_genes <- c("XCR1", "IRF8", "CD74", "CD52", "CLEC9A", "TAP1")
cdc2_genes <- c("CD14", "CD163", "ITGAM", "SIRPA", "CD2", "CD1C", "CX3CR1")
# cytokines_and_chemokines <- c("CXCL8", "CXCL2")
dc_activation <- c("CCR7")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "DC") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% hypoxia_genes ~ "hypoxia/angiogenesis" ,
    gene %in% cdc1_genes ~ "cDC1" ,
    gene %in% cdc2_genes ~ "cDC2" ,
    gene %in% dc_activation ~ "mDC" ,
    # gene %in% cytokines_and_chemokines ~ "cytokines and chemokines" ,
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            hjust = 0) + 
  labs(title = "DC")
```

```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
immunosuppressive_genes <- c("CD163", "SPP1", "CCL2", "CSF1", "CSF1R", "ARG1")
proinflammatory_genes <- c("IL6", "C1QB", "NOS2", "TLR2", "TLR4", "TNF", "IL1B")
adhesion_genes <- c("VCAM1")
cytokines_and_chemokines <- c("IL33", "CXCL6", "CXCL10", "CXCL11", "CXCL8", "CXCL2")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "TAM") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    # gene %in% hypoxia_genes ~ "hypoxia response" ,
    gene %in% immunosuppressive_genes ~ "immunosuppressive",
    gene %in% proinflammatory_genes ~ "proinflammatory",
    gene %in% adhesion_genes ~ "adhesion",
    gene %in% cytokines_and_chemokines ~ "cytokines and chemokines",
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            hjust = 1, 
            max.overlaps = Inf) + 
  labs(title = "TAM")
```


```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
angiogenesis_genes <- c("SERPINE1", "MMP1", "ANGPT2", "PDGFA", "THBS1", "ESM1", "VWF")
cytokines_and_chemokines <- c("IL33", "CXCL6", "CXCL10")
adhesion <- c("EPCAM", "VCAM1")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "Endothelial cells") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    gene %in% angiogenesis_genes ~ "angiogenesis" ,
    gene %in% cytokines_and_chemokines ~ "cytokines and chemokines", 
    gene %in% adhesion ~ "adhesion", 
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            force = 5, 
            hjust = 1) + 
  labs(title = "Endothelial cells")
```

```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
cytokines_and_chemokines <- c("IL6", "IL33", "CXCL6", "CXCL8")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "Fibroblasts") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    gene %in% cytokines_and_chemokines ~ "cytokines and chemokines", 
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            hjust = 1) + 
  labs(title = "Fibroblasts")
```


```{r}
hypoxia_genes <- c("HIF1A", "EPAS1", "VEGFA")
stromal_crosstalk <- c("FN1")
cancer_genes <- c("ANGPTL4")
cytokines_and_chemokines <- c("CXCL8", "CXCL2")

model_out %>% 
  ungroup() %>% 
  filter(cell_type == "Tumor") %>% 
  filter(term == "mean_hypoxia") %>% 
  arrange(desc(estimate)) %>% 
  mutate(category = case_when(
    gene %in% hypoxia_genes ~ "hypoxia response" ,
    gene %in% stromal_crosstalk ~ "stromal crosstalk",
    gene %in% cytokines_and_chemokines ~ "cytokines and chemokines", 
    gene %in% cancer_genes ~ "invasiveness", 
    TRUE ~ NA_character_
  )) %>% 
  arrange(estimate) %>% 
  mutate(rank = row_number()) %>% 
  ggplot() + 
  aes(rank, estimate) + 
  geom_hline(yintercept = 0, linetype = "dotted") + 
  geom_line() + 
  geom_segment(aes(x = rank, xend = rank, 
                   y = 0, yend = estimate, 
                   color = category, 
                   linetype = p.value < 0.1), 
               data = ~..1 %>% drop_na(category)) + 
  geom_point(aes(color = category), 
             data = ~..1 %>% drop_na(category), size = 3) + 
  geom_text_repel(aes(label = gene, color = category), 
            data = ~..1 %>% drop_na(category), 
            hjust = 1) + 
  labs(title = "Tumor")
```

# Limma-voom

## Pre-processing

```{r}
pb_df <- c(pb_imm_list, pb_stromal_list, list(Tumor = pb_tumor)) %>% 
  map_dfr(as_tibble, rownames = "sample_ID", .id = "cell_type") %>% 
  separate(sample_ID, sep = "_", into = c("sample_ID", "hypoxia_hi.v.lo"))

pb_tidy_lv <- pb_df %>% 
  pivot_longer(cols = 4:ncol(.), 
               names_to = "gene", 
               values_to = "count")

pb_mtx_list <- pb_tidy_lv %>%
  ungroup() %>%
  named_group_split(cell_type) %>% 
  map(~{
    ..1 %>% 
      drop_na(count) %>% 
      mutate(sample_celltype = paste0(sample_ID, "__", cell_type)) %>% 
      pivot_wider(
        id_cols = gene,
        names_from = sample_celltype,
        values_from = count, 
        values_fill = 0
      ) %>%
      column_to_rownames("gene") %>%
      as.matrix()
  })
```

```{r}
d0_list <- map(pb_mtx_list, DGEList)
d0_list <- map(d0_list, calcNormFactors)
```

```{r}
fit_list <- map(
  d0_list, 
  ~{
    pb_metadata_df <- 
      tibble(sample_celltype = colnames(..1)) %>% 
      separate(sample_celltype, 
               into = c("sample_ID", "cell_type"), 
               sep = "__", 
               remove = FALSE) %>% 
      left_join(dplyr::select(mean_hypoxia_df, sample_ID, mean_hypoxia), 
                by = "sample_ID")
    
    mm <- model.matrix(~0 + mean_hypoxia, data = pb_metadata_df)
    y <- voom(..1, mm, plot = T)
    fit <- lmFit(y, mm)
  }
)
```

```{r}
fit_df <- fit_list %>% 
  map_dfr(~as_tibble(..1$coefficients, rownames = "gene"), 
          .id = "cell_type")
```

## Cross with model

```{r, fig.asp = 1}
model_out %>% 
  filter(term == "mean_hypoxia") %>% 
  left_join(fit_df, by = c("cell_type", "gene")) %>% 
  mutate(mean_hypoxia = ifelse(is.na(mean_hypoxia), -Inf, mean_hypoxia)) %>% 
  ggplot() + 
  aes(estimate, mean_hypoxia) + 
  geom_point() + 
  geom_text_repel(aes(label = gene), 
                  data = ~..1 %>% filter(gene %in% c(
                    "ACTB"
                  )), 
                  color = "blue") + 
  facet_wrap(~cell_type, ncol = 3) + 
  labs(x = "estimate from linear model", 
       y = "estimate from limma-voom")
```

```{r, fig.asp = 1}
model_out %>% 
  filter(term == "mean_hypoxia") %>% 
  left_join(fit_df, by = c("cell_type", "gene")) %>% 
  mutate(mean_hypoxia = ifelse(is.na(mean_hypoxia), -Inf, mean_hypoxia)) %>% 
  ggplot() + 
  aes(estimate, mean_hypoxia) + 
  geom_density_2d() + 
  facet_wrap(~cell_type, ncol = 3) + 
  labs(x = "estimate from linear model", 
       y = "estimate from limma-voom")
```

# Spearman cor

```{r}
genes_subset <- model_out$gene %>% unique()
spearman_df <- run_or_readRDS(
  file.path(file_cache_dir, "spearman.rds"), 
  function() {
    spearman_df <- pb_tidy %>%
      filter(gene %in% genes_subset) %>% 
      left_join(mean_hypoxia_df, by = c("sample_ID")) %>%
      group_by(cell_type, gene) %>%
      summarise(spearman_cor = cor(log1p_CPM, mean_hypoxia.x, method = "spearman"))
  }
)
```

```{r, fig.asp = 1}
spearman_df %>% 
  drop_na(spearman_cor) %>% 
  left_join(filter(model_out, term == "mean_hypoxia"), 
            by = c("cell_type", "gene")) %>% 
  drop_na(estimate) %>% 
  ggplot() + 
  aes(estimate, spearman_cor) + 
  geom_point() + 
  facet_wrap(~cell_type, ncol = 3)
```

# Comparison bw LM, Limma-voom, and spearman cor

# Expression of vimentin (VIM)

```{r}
sc_pb_tidy <- readRDS(here("results", "sc_pb_tidy.rds"))
```

```{r}
plot_gene_expr_from_pb <- function(gene_str) {
  
  sc_pb_tidy %>%
    filter(gene == gene_str) %>%
    filter(!(
      cell_type %in% c("Stressed", "Contaminating", "Doublets", "Neutrophil")
    )) %>%
    ggplot() +
    aes(cell_type, log1p_CPM) +
    geom_col(
      aes(y = mean_CPM, fill = cell_type),
      data = ~ ..1 %>% group_by(cell_type) %>% summarise(mean_CPM = mean(log1p_CPM))
    ) +
    geom_jitter(width = 0.2, size = 2) +
    labs(y = sprintf("%s expression", gene_str)) +
    scale_fill_manual(values = c(palette_specific_immune, palette_general_cell_type)) +
    theme(
      legend.position = "none",
      axis.ticks.y = element_blank(),
      axis.line.y = element_blank()
    ) +
    scale_y_continuous(expand = c(0, 0), limits = ~ c(..1[1], ..1[2] * 1.1)) +
    coord_flip()
  
}
```

```{r}
plot_gene_expr_from_pb("VIM")
plot_gene_expr_from_pb("PTPRC")
plot_gene_expr_from_pb("CD8A")
plot_gene_expr_from_pb("CD3D")
plot_gene_expr_from_pb("MLANA")
plot_gene_expr_from_pb("PECAM1")
plot_gene_expr_from_pb("FCN3")
```


# sessionInfo

```{r}
sessionInfo()
```

