---
title: "CellPhoneDB Hypoxia Comparison"
execute: 
  freeze: auto
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
source(here("lib", "single_cell", "helper_Seurat.R"))

# document specific packages: 
library(Seurat)

file_cache_dir <- file.path(cache_dir, "sc-processing-cellphonedb")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

```{r}
mean_hypoxia_df <- readRDS(file.path(data_dir, "scRNAseq", "mean_hypoxia_df.rds"))
cpdb_tidy <- readRDS(file.path(results_dir, "cpdb_tidy.rds")) %>% 
    mutate(sample_ID = str_remove(id, "cpdb_")) %>%
    left_join(mean_hypoxia_df, by = "sample_ID") 
```

# Plot specific interactions

```{r, fig.asp = 1}
plot_interacting_pair <- function(cpdb_tidy, interacting_pair_str, 
                                  cell_type_str) {
  cpdb_tidy %>%
    filter(interacting_pair == interacting_pair_str) %>%
    filter(str_detect(name, cell_type_str)) %>%
    drop_na(value.mean) %>%
    ggplot() +
    aes(id, name, size = value.mean) +
    geom_point(aes(color = -log10(value.pval + 1e-4))) +
    geom_point(
      shape = 1,
      color = "red",
      stroke = 2,
      data = ~ ..1 %>%
        filter(value.pval < 0.05)
    ) +
    scale_color_viridis() +
    rotate_x_text() +
    labs(title = interacting_pair_str) +
    theme(legend.position = "bottom")
}
```

```{r, fig.asp = 1}
plot_interacting_pair(cpdb_tidy, "IL10_receptor_IL10", "DC") + 
  facet_grid(~hypoxia_hi.v.lo, scale = "free_x")
```

# Chord Diagrams

## Example

```{r}
data <- matrix(c(
  10, 30, 40, 20,
  25, 15, 35, 30,
  15, 25, 20, 10,
  30, 20, 10, 15
), ncol = 4, byrow = TRUE)

# Create a chord diagram with annotations
chordDiagram(
  x = data,
  transparency = 0.5, 
  directional = 1
)
```

## Data Set up

```{r}
cpdb_tidy_sep <- cpdb_tidy %>% 
  separate(name, into = c("from", "to"), sep = "\\|")
```

## Test visualizations

```{r}
data <- cpdb_tidy_sep %>%
  filter(id == "cpdb_11967") %>%
  filter(annotation_strategy == "curated") %>% 
  mutate(signif = case_when(value.pval %>% is.na() ~ FALSE,
                            value.pval < 0.05 ~ TRUE,
                            TRUE ~ FALSE)) %>%
  group_by(from, to, signif) %>%
  summarise(value = n()) %>%
  mutate(color = ifelse(signif, "black", NA))
chordDiagram(
  x = data %>% select(-signif),
  transparency = 0.5,
  col = data$color, 
  directional = 1,
  direction.type = "arrows",
  link.arr.length = 0.2,
  link.lwd = 1, 
  link.lty = 1, 
  link.border = "black"
)
```


```{r}
data <- cpdb_tidy_sep %>%
  filter(id == "cpdb_11967") %>%
  filter(annotation_strategy == "curated") %>% 
  mutate(signif = case_when(value.pval %>% is.na() ~ FALSE,
                            value.pval < 0.05 ~ TRUE,
                            TRUE ~ FALSE)) %>%
  group_by(from, to) %>%
  summarise(value = sum(signif) > 0) %>% 
  mutate(color = ifelse(value, "grey", NA))
chordDiagram(
  x = data,
  transparency = 0.5,
  col = data$color, 
  directional = 0
)
```

## For each sample

### Hypoxic samples

```{r}
id_strs <- cpdb_tidy_sep %>% 
  filter(hypoxia_hi.v.lo == "high hypoxia") %>% 
  .$id %>% 
  unique()
for (id_str in id_strs) {
  data <- cpdb_tidy_sep %>% 
    filter(id == id_str) %>% 
    filter(annotation_strategy == "curated") %>% 
    mutate(signif = case_when(
      value.pval %>% is.na() ~ FALSE, 
      value.pval < 0.05 ~ TRUE, 
      TRUE ~ FALSE
    )) %>% 
    group_by(from, to, signif) %>% 
    summarise(count = n()) %>% 
    mutate(color = ifelse(signif, "red", NA))
  chordDiagram(
    x = data %>% select(-signif),
    col = data$color, 
    directional = 1,
    direction.type = "arrows", 
    link.arr.length = 0.2,
    link.lwd = 1, 
    link.lty = 1, 
    link.border = "black"
  )
  text(0,0,str_remove(id_str, "cpdb_"))
  circos.clear()
}
```

### Non-hypoxic samples

```{r}
id_strs <- cpdb_tidy_sep %>% 
  filter(hypoxia_hi.v.lo == "low hypoxia") %>% 
  .$id %>% 
  unique()
for (id_str in id_strs) {
  data <- cpdb_tidy_sep %>% 
    filter(id == id_str) %>% 
    filter(annotation_strategy == "curated") %>% 
    mutate(signif = case_when(
      value.pval %>% is.na() ~ FALSE, 
      value.pval < 0.05 ~ TRUE, 
      TRUE ~ FALSE
    )) %>% 
    group_by(from, to, signif) %>% 
    summarise(count = n()) %>% 
    mutate(color = ifelse(signif, "red", NA))
  chordDiagram(
    x = data %>% select(-signif),
    col = data$color, 
    directional = 1,
    direction.type = "arrows", 
    link.arr.length = 0.2,
    link.lwd = 1, 
    link.lty = 1, 
    link.border = "black"
  )
  text(0,0,str_remove(id_str, "cpdb_"))
  circos.clear()
}
```


### Mid hypoxic

```{r}
id_strs <- cpdb_tidy_sep %>% 
  filter(hypoxia_hi.v.lo == "mid hypoxia") %>% 
  .$id %>% 
  unique()
for (id_str in id_strs) {
  data <- cpdb_tidy_sep %>% 
    filter(id == id_str) %>% 
    filter(annotation_strategy == "curated") %>% 
    mutate(signif = case_when(
      value.pval %>% is.na() ~ FALSE, 
      value.pval < 0.05 ~ TRUE, 
      TRUE ~ FALSE
    )) %>% 
    group_by(from, to, signif) %>% 
    summarise(count = n()) %>% 
    mutate(color = ifelse(signif, "red", NA))
  chordDiagram(
    x = data %>% select(-signif),
    col = data$color, 
    directional = 1,
    direction.type = "arrows", 
    link.arr.length = 0.2,
    link.lwd = 1, 
    link.lty = 1, 
    link.border = "black"
  )
  text(0,0,str_remove(id_str, "cpdb_"))
  circos.clear()
}
```

# Number of interactions, total

```{r}
n_int_df <- cpdb_tidy_sep %>% 
  mutate(value.qval = p.adjust(value.pval, method = "BH")) %>% 
  filter(value.qval < 0.05) %>% 
  group_by(sample_ID, from, to) %>% 
  summarise(n_int = n(), 
            median_interaction_strength = median(value.mean), 
            mean_interaction_strength = mean(value.mean)) %>% 
  group_by(from, to) %>% 
  mutate(mean_n_int = mean(n_int)) %>% 
  ungroup()
```

```{r, fig.asp = 0.8, fig.width = 5}
n_int_df %>% 
  ggplot() + 
  aes(1, n_int, fill = mean_n_int) + 
  geom_violin(scale = "width", linewidth = 0.1) + 
  facet_grid(factor(from, levels = sort(unique(from)))~factor(to, levels = rev(sort(unique(to))))) + 
  scale_fill_viridis() + 
  theme(axis.text = element_blank(), 
        strip.background = element_blank(), 
        strip.text.y = element_text(angle = 0, hjust = 0), 
        strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5)) + 
  remove_spines()
```

# Number of interactions, w hypoxia

## Stromal vs tumor

The number of stromal-tumor interactions increases as hypoxia increases, across all immune states. 

```{r}
n_int_stromal <- cpdb_tidy_sep %>% 
  filter((from == "Stromal" & to == "Tumor") | (to == "Stromal" & from == "Tumor")) %>% 
  mutate(value.qval = p.adjust(value.pval, method = "BH")) %>% 
  filter(value.qval < 0.05) %>% 
  group_by(sample_ID) %>% 
  summarise(ninteractions = n(), 
            median_interaction_strength = median(value.mean)) %>% 
  left_join(mean_hypoxia_df, by = "sample_ID")
```

```{r}
n_int_stromal %>% 
  # filter(immune_hi.v.lo == "immune high") %>% 
  ggplot() + 
  aes(mean_hypoxia, ninteractions) + 
  geom_point(size = 3) + 
  stat_smooth(method = "lm") + 
  stat_cor() + 
  facet_grid(~immune_hi.v.lo)
```

## Stromal vs myeloid

The number of stromal-myeloid interactions is not associated with hypoxia in immune high

```{r}
n_int_stromal_m <- cpdb_tidy_sep %>% 
  filter((from == "Stromal" & to == "DC") | (to == "Stromal" & from == "DC") | (from == "Stromal" & to == "Mac/mono") | (to == "Stromal" & from == "Mac/mono") | (from == "Stromal" & to == "Neutrophil") | (to == "Stromal" & from == "Neutrophil")) %>% 
  mutate(value.qval = p.adjust(value.pval, method = "BH")) %>% 
  filter(value.qval < 0.05) %>% 
  group_by(sample_ID) %>% 
  summarise(ninteractions = n(), 
            median_interaction_strength = median(value.mean)) %>% 
  left_join(mean_hypoxia_df, by = "sample_ID")
```

```{r}
n_int_stromal_m %>% 
  # filter(immune_hi.v.lo == "immune high") %>% 
  ggplot() + 
  aes(mean_hypoxia, ninteractions) + 
  geom_point(size = 3) + 
  stat_smooth(method = "lm") + 
  stat_cor() + 
  facet_grid(~immune_hi.v.lo)
```

## Stromal vs immune

The number of stromal-immune interactions is not associated with hypoxia in immune high

```{r}
n_int_stromal_imm <- cpdb_tidy_sep %>% 
  filter(from == "Stromal" | to == "Stromal") %>% 
  filter(!(from == "Stromal" & to == "Tumor")) %>% 
  filter(!((to == "Stromal" & from == "Tumor"))) %>% 
  mutate(value.qval = p.adjust(value.pval, method = "BH")) %>% 
  filter(value.qval < 0.05) %>% 
  group_by(sample_ID) %>% 
  summarise(ninteractions = n(), 
            median_interaction_strength = median(value.mean)) %>% 
  left_join(mean_hypoxia_df, by = "sample_ID")
```

```{r}
n_int_stromal_imm %>% 
  # filter(immune_hi.v.lo == "immune high") %>% 
  ggplot() + 
  aes(mean_hypoxia, ninteractions) + 
  geom_point(size = 3) + 
  stat_smooth(method = "lm") + 
  stat_cor() + 
  facet_grid(~immune_hi.v.lo)
```

# What stromal to tumor interactions

```{r}
cpdb_stromal_tumor <- cpdb_tidy_sep %>% 
  filter((from == "Stromal" & to == "Tumor") | (to == "Stromal" & from == "Tumor")) %>% 
  mutate(value.qval = p.adjust(value.pval, method = "BH")) %>% 
  filter(value.qval < 0.05) 
```

```{r}
cpdb_stromal_tumor %>% 
  filter(immune_hi.v.lo == "immune high") %>% 
  group_by(interacting_pair, from, to) %>% 
  summarise(n = n(),
            avg.value.mean = mean(value.mean)) %>%
  arrange(desc(n), desc(avg.value.mean)) %>% 
  filter(n > 3) %>% 
  mutate(interacting_pair_from_to = paste0(interacting_pair, "_from", from, "_to", to)) %>% 
  ggplot() + 
  aes(fct_reorder(interacting_pair_from_to, n), n) + 
  geom_col() + 
  coord_flip()
```


```{r}
cpdb_stromal_tumor %>% 
  filter(immune_hi.v.lo == "immune high") %>% 
  group_by(interacting_pair, from, to, gene_a, gene_b) %>% 
  summarise(n = n(),
            avg.value.mean = mean(value.mean)) %>%
  arrange(desc(n), desc(avg.value.mean)) %>% 
  head(10)
```

```{r}
cpdb_stromal_tumor %>% 
  filter(immune_hi.v.lo == "immune low") %>% 
  group_by(interacting_pair, from, to, gene_a, gene_b) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n)) %>% 
  head(10)
```

# What interactions with stromal cells

```{r}
cpdb_tidy_sep %>% 
  filter(from == "Stromal" | to == "Stromal") %>% 
  filter(value.pval < 0.05) %>% 
  group_by(interacting_pair, hypoxia_hi.v.lo) %>% 
  summarise(n_samples = length(unique(sample_ID))) %>% 
  arrange(desc(n_samples)) %>% 
  pivot_wider(id_cols = interacting_pair, 
              names_from = hypoxia_hi.v.lo, 
              values_from = n_samples, 
              values_fill = 0) %>% 
  mutate(rel_high_vs_low = `high hypoxia` - `low hypoxia`) %>% 
  arrange(desc(rel_high_vs_low))
```

```{r}
cpdb_tidy_sep %>% 
  filter(from == "Stromal" | to == "Stromal") %>% 
  filter(from == "DC" | to == "DC") %>% 
  filter(value.pval < 0.05) %>% 
  filter(annotation_strategy == "curated") %>% 
  arrange(desc(value.mean))
```



