---
title: "Bulk RNA-seq Clinical Features"
execute: 
  freeze: auto
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
```

# Purpose

The purpose of this chapter is to lay out some of the clinical characteristics in the bulk cohort that we're analyzing. 

# Set up

We'll start by reading in the clinical metadata table.  

```{r}
bulk_clin_metadata <- read_tsv(here("data", "bulk", "bulk-clin-metadata.tsv"))
```

## Data Cleaning

```{r}
bulk_clin_metadata <- bulk_clin_metadata %>% 
  mutate(age = as.numeric(age))
```


# Characteristics

Our bulk cohort is made up of 8 previously published cohorts:

```{r}
bulk_clin_metadata$cohort %>% unique()
```

In our metadata table, we've collected the following features:  

```{r}
colnames(bulk_clin_metadata)
```

In total, there are 693 patients in our aggregated dataset. 

```{r}
nrow(bulk_clin_metadata)
```

## Age

```{r}
p <- bulk_clin_metadata %>%
  ggplot() +
  aes(age) +
  geom_histogram(bins = 10) +
  remove_x_spine() + 
  scale_y_continuous(expand = c(0, 0)) 
p + 
  geom_segment(aes(x = median(age), xend = median(age), y = 0, yend = Inf), 
               linetype = "dashed") + 
  geom_text(aes(x = median(age), y = 60, label = paste0("median=", median(age))), 
            hjust = 1.1) + 
  labs(y = "Freq. ")
```

```{r}
median_age_df <-
  bulk_clin_metadata %>% 
  group_by(group) %>% 
  summarise(median_age = median(age))
p +
  facet_grid(group ~ .) +
  geom_segment(
    aes(
      x = median_age,
      xend = median_age,
      y = 0,
      yend = Inf
    ),
    linetype = "dashed",
    data = median_age_df
  ) +
  geom_text(
    aes(
      x = median_age,
      y = 60,
      label = paste0("median=", median_age)
    ),
    hjust = 1.1,
    data = median_age_df
  ) 
```

```{r, fig.asp = 1}
#| fig-asp: 1

median_age_df <-
  bulk_clin_metadata %>% 
  group_by(cohort) %>% 
  summarise(median_age = median(age))
p +
  facet_grid(cohort ~ .) +
  geom_segment(
    aes(
      x = median_age,
      xend = median_age,
      y = 0,
      yend = Inf
    ),
    linetype = "dashed",
    data = median_age_df
  ) +
  geom_text(
    aes(
      x = median_age,
      y = 30,
      label = paste0("median=", median_age)
    ),
    hjust = 1.1,
    data = median_age_df
  ) + 
  theme(strip.text.y = element_text(angle = 0))
```

## Sex

```{r, fig.width = 3, fig.asp = 1}
#| fig-width: 3
#| fig-asp: 1

count_df <- bulk_clin_metadata %>% 
  group_by(sex) %>% 
  count() %>% 
  ungroup() %>%
  mutate(pct = n / sum(n))
count_df %>% 
  ggplot() + 
  aes(1, n) + 
  geom_col(aes(fill = sex)) + 
  annotate("text", x = 1, y = count_df$n[[2]], vjust = 1.3, 
           label = paste0(format(count_df$pct[[2]]*100, digits = 3), "%")) + 
  scale_y_continuous(expand = c(0, 0)) + 
  scale_fill_manual(values = palette_sex) + 
  remove_x_spine() + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) + 
  labs(y = "Freq. ")
```

```{r, fig.width = 3, fig.asp = 1}
#| fig-width: 3
#| fig-asp: 1

count_df <- bulk_clin_metadata %>% 
  group_by(sex, group) %>% 
  count() %>% 
  group_by(group) %>% 
  arrange(desc(sex)) %>% 
  mutate(pct = n / sum(n)) %>% 
  mutate(cumsum_pct = cumsum(pct)) %>% 
  ungroup()
count_df %>% 
  ggplot() + 
  aes(1, pct) + 
  geom_col(aes(fill = sex)) + 
  geom_text(aes(y = cumsum_pct, 
                label = paste0(format(pct*100, digits = 3), "%")), 
            vjust = 1, 
            size = 10*GGPLOT_TEXT_SCALE_FACTOR) + 
  facet_grid(~group) + 
  scale_y_continuous(expand = c(0, 0)) + 
  scale_fill_manual(values = palette_sex) + 
  remove_x_spine() + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) + 
  labs(y = "Freq. ")
```

## Tumor type

```{r, fig.width = 3, fig.height = 3}
#| fig-width: 3
#| fig-height: 3

count_df <- bulk_clin_metadata %>% 
  mutate(tumor_type = case_when(
    is.na(tumor_type) | tumor_type %in% c("cutaneous and eye", "muco-cutaneous") ~ "unknown/other", 
    TRUE ~ tumor_type
  )) %>% 
  group_by(tumor_type) %>% 
  count() %>%
  ungroup() %>% 
  arrange(desc(tumor_type)) %>% 
  mutate(pct = n / sum(n)) %>% 
  mutate(cumsum_n = cumsum(n))
count_df %>% 
  ggplot() + 
  aes(1, n) + 
  geom_col(aes(fill = tumor_type)) + 
  geom_text(aes(y = cumsum_n, 
                label = paste0(format(pct*100, digits = 3), "%")), 
            vjust = 1, 
            size = 10*GGPLOT_TEXT_SCALE_FACTOR) + 
  scale_y_continuous(expand = c(0, 0)) + 
  remove_x_spine() + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) + 
  labs(y = "Freq. ")
```


```{r, fig.width = 4, fig.height = 3}
#| fig-width: 4
#| fig-height: 3

count_df <- bulk_clin_metadata %>% 
  mutate(tumor_type = case_when(
    is.na(tumor_type) | tumor_type %in% c("cutaneous and eye", "muco-cutaneous") ~ "unknown/other", 
    TRUE ~ tumor_type
  )) %>% 
  group_by(tumor_type, group) %>% 
  count() %>%
  group_by(group) %>% 
  arrange(desc(tumor_type)) %>% 
  mutate(pct = n / sum(n)) %>% 
  mutate(cumsum_pct = cumsum(pct)) %>% 
  ungroup()
count_df %>% 
  ggplot() + 
  aes(1, pct) + 
  geom_col(aes(fill = tumor_type)) + 
  geom_text(aes(y = cumsum_pct, 
                label = paste0(format(pct*100, digits = 3), "%")), 
            vjust = 1, 
            size = 10*GGPLOT_TEXT_SCALE_FACTOR) + 
  facet_grid(~group) + 
  scale_y_continuous(expand = c(0, 0)) + 
  remove_x_spine() + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) + 
  labs(y = "Freq. ")
```

## prior IO

## prior therapy

## treatment

```{r, fig.width = 4, fig.height = 3}
#| fig-width: 4
#| fig-height: 3

count_df <- bulk_clin_metadata %>% 
  group_by(treatment) %>% 
  count() %>%
  ungroup() %>% 
  arrange(desc(treatment)) %>% 
  mutate(pct = n / sum(n)) %>% 
  mutate(cumsum_n = cumsum(n))
count_df %>% 
  ggplot() + 
  aes(1, n) + 
  geom_col(aes(fill = treatment)) + 
  geom_text(aes(y = cumsum_n, 
                label = paste0(format(pct*100, digits = 3), "%")), 
            vjust = 1, 
            size = 10*GGPLOT_TEXT_SCALE_FACTOR) + 
  scale_y_continuous(expand = c(0, 0)) + 
  remove_x_spine() + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) + 
  labs(y = "Freq. ")
```

```{r, fig.width = 4, fig.height = 3}
#| fig-width: 4
#| fig-height: 3

count_df <- bulk_clin_metadata %>% 
  group_by(treatment, group) %>% 
  count() %>%
  group_by(group) %>% 
  arrange(desc(treatment)) %>% 
  mutate(pct = n / sum(n)) %>% 
  mutate(cumsum_pct = cumsum(pct)) %>% 
  ungroup()
count_df %>% 
  ggplot() + 
  aes(1, pct) + 
  geom_col(aes(fill = treatment)) + 
  geom_text(aes(y = cumsum_pct, 
                label = paste0(format(pct*100, digits = 3), "%")), 
            vjust = 1, 
            size = 10*GGPLOT_TEXT_SCALE_FACTOR) + 
  facet_grid(~group) + 
  scale_y_continuous(expand = c(0, 0)) + 
  remove_x_spine() + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) + 
  labs(y = "Freq. ")
```

## response

```{r, fig.width = 3, fig.height = 3}
#| fig-width: 3
#| fig-height: 3

count_df <- bulk_clin_metadata %>% 
  mutate(response_splitSD = ifelse(is.na(response_splitSD), "NA", response_splitSD)) %>% 
  group_by(response_splitSD) %>% 
  count() %>%
  ungroup() %>% 
  arrange(desc(response_splitSD)) %>% 
  mutate(pct = n / sum(n)) %>% 
  mutate(cumsum_n = cumsum(n))
count_df %>% 
  ggplot() + 
  aes(1, n) + 
  geom_col(aes(fill = response_splitSD)) + 
  geom_text(aes(y = cumsum_n, 
                label = paste0(format(pct*100, digits = 3), "%")), 
            vjust = 1, 
            size = 10*GGPLOT_TEXT_SCALE_FACTOR) + 
  scale_fill_manual(values = palette_response) + 
  scale_y_continuous(expand = c(0, 0)) + 
  remove_x_spine() + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) + 
  labs(y = "Freq. ")
```

```{r, fig.width = 4, fig.height = 3}
#| fig-width: 4
#| fig-height: 3

count_df <- bulk_clin_metadata %>% 
  mutate(response_splitSD = ifelse(is.na(response_splitSD), "NA", response_splitSD)) %>% 
  group_by(response_splitSD, group) %>% 
  count() %>%
  group_by(group) %>% 
  arrange(desc(response_splitSD)) %>% 
  mutate(pct = n / sum(n)) %>% 
  mutate(cumsum_pct = cumsum(pct)) %>% 
  ungroup()
count_df %>% 
  ggplot() + 
  aes(1, pct) + 
  geom_col(aes(fill = response_splitSD)) + 
  geom_text(aes(y = cumsum_pct, 
                label = paste0(format(pct*100, digits = 3), "%")), 
            vjust = 1, 
            size = 10*GGPLOT_TEXT_SCALE_FACTOR) + 
  scale_fill_manual(values = palette_response) + 
  facet_grid(~group) + 
  scale_y_continuous(expand = c(0, 0)) + 
  remove_x_spine() + 
  theme(axis.text.x = element_blank(), 
        axis.title.x = element_blank()) + 
  labs(y = "Freq. ")
```

# Response and Progression

## Discovery cohort

```{r}
bulk_clin_metadata %>% 
  filter(group == "discovery") %>% 
  drop_na(response_splitSD) %>% 
  filter(is.na(excluded_clin)) %>% 
  mutate(progressed = case_when(
    is.na(censor_pfs_event_1_progressed_0) ~ "unknown",
    censor_pfs_event_1_progressed_0 == 0 ~ "progressed", 
    censor_pfs_event_1_progressed_0 == 1 ~ "did not progress", 
    TRUE ~ NA_character_
  )) %>% 
  mutate(response = ifelse(is.na(response), "SD", response)) %>% 
  group_by(treatment, response, progressed) %>% 
  count() %>% 
  pivot_wider(id_cols = c(treatment, response), 
              names_from = progressed, 
              values_from = n, 
              values_fill = 0) %>% 
  mutate(progressed_vs_not = progressed / (`did not progress` + progressed)) %>% 
  relocate(progressed_vs_not, .before = `did not progress`)
```

```{r}

count_treatment_imm_progressed_df <- bulk_clin_metadata %>% 
  filter(group == "discovery") %>% 
  filter(is.na(excluded_clin)) %>% 
  mutate(progressed = case_when(
    is.na(censor_pfs_event_1_progressed_0) ~ "unknown",
    censor_pfs_event_1_progressed_0 == 0 ~ "progressed", 
    censor_pfs_event_1_progressed_0 == 1 ~ "did not progress", 
    TRUE ~ NA_character_
  )) %>% 
  group_by(treatment, progressed, immune_hi.v.lo) %>% 
  count() 
count_treatment_imm_progressed_df %>% 
  pivot_wider(id_cols = c(treatment, immune_hi.v.lo), 
              names_from = progressed, 
              values_from = n, 
              values_fill = 0) %>% 
  mutate(progressed_vs_not = progressed / (`did not progress` + progressed)) %>% 
  relocate(progressed_vs_not, .before = `did not progress`)

count_treatment_imm_progressed_df %>% 
  pivot_wider(id_cols = c(treatment, progressed), 
              names_from = immune_hi.v.lo, 
              values_from = n, 
              values_fill = 0) %>% 
  mutate(immune_high_vs_not = `immune high` / (`immune high` + `immune low`)) %>% 
  relocate(immune_high_vs_not, .before = `immune high`)
```


