---
title: "CellPhoneDB Processing"
execute: 
  freeze: auto
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
source(here("lib", "single_cell", "helper_Seurat.R"))

# document specific packages: 
library(Seurat)

file_cache_dir <- file.path(cache_dir, "sc-processing-cellphonedb")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

# Cell type specific

## Read in

```{r}
cpdb_subdirs <- list.files(file.path(results_dir, "cpdb")) %>% 
  str_subset("cpdb") %>% 
  str_subset("10x_dir", negate = T)

cpdb_compartment_means <- cpdb_subdirs %>% 
  set_names() %>% 
  map_dfr(~read_tsv(file.path(results_dir, "cpdb", ..1, "out", "cell_type_specific", "means.txt")), .id = "id")

cpdb_compartment_pvals <- cpdb_subdirs %>% 
  set_names() %>% 
  map_dfr(~read_tsv(file.path(results_dir, "cpdb", ..1, "out", "cell_type_specific", "pvalues.txt")), .id = "id")

cpdb_means_tidy <- cpdb_compartment_means %>% 
  pivot_longer(cols = str_subset(colnames(.), "\\|"))
cpdb_pvals_tidy <- cpdb_compartment_pvals %>% 
  pivot_longer(cols = str_subset(colnames(.), "\\|"))
cpdb_tidy <- cpdb_means_tidy %>%
  left_join(
    cpdb_pvals_tidy,
    by = join_by(
      id,
      id_cp_interaction,
      interacting_pair,
      partner_a,
      partner_b,
      gene_a,
      gene_b,
      secreted,
      receptor_a,
      receptor_b,
      annotation_strategy,
      is_integrin,
      name
    ), 
    suffix = c(".mean", ".pval")
  )
```

## Export/save

```{r}
saveRDS(cpdb_tidy, file.path(results_dir, "cpdb_tidy.rds"))
```

# Properties

```{r}
cpdb_tidy$id_cp_interaction %>% unique() %>% length()
cpdb_tidy$interacting_pair %>% unique() %>% length()
```

# Summarize total interactions

For one sample

```{r}
summarize_total_ints <- function(cpdb_tidy, id_str) {
  
  h_mtx <- cpdb_tidy %>% 
    filter(id == id_str) %>% 
    mutate(significant = ifelse(is.na(value.pval), 
                                FALSE, 
                                value.pval < 0.05)) %>% 
    group_by(name, significant) %>% 
    count() %>% 
    pivot_wider(id_cols = name, names_from = significant, values_from = n, 
                values_fill = 0) %>% 
    separate(name, into = c("c1", "c2"), sep = "\\|") %>% 
    mutate(prop_sig = `TRUE` / (`TRUE` + `FALSE`) * 100) %>% 
    pivot_wider(id_cols = c1, 
                names_from = c2, 
                values_from = prop_sig) %>% 
    column_to_rownames("c1") %>% 
    as.matrix() 
  ht <- Heatmap(
    h_mtx, 
    col = colorRamp2(breaks = c(0, 10), 
                     colors = palette_cor[-1]), 
    name = "% signif. interaction", 
    width = unit(2, "in"), 
    height = unit(2, "in"), 
    column_title = id_str
  )
  return(draw(ht))
}

```

```{r}
summarize_total_ints(cpdb_tidy, "cpdb_9129")
```


# Plot specific interactions

```{r, fig.asp = 1}
plot_interacting_pair <- function(cpdb_tidy, interacting_pair_str, 
                                  cell_type_str) {
  cpdb_tidy %>%
    filter(interacting_pair == interacting_pair_str) %>%
    filter(str_detect(name, cell_type_str)) %>%
    drop_na(value.mean) %>%
    mutate(sample_ID = str_remove(id, "cpdb_")) %>%
    ggplot() +
    aes(id, name, size = value.mean) +
    geom_point(aes(color = -log10(value.pval + 1e-4))) +
    geom_point(
      shape = 1,
      color = "red",
      stroke = 2,
      data = ~ ..1 %>%
        filter(value.pval < 0.05)
    ) +
    scale_color_viridis() +
    rotate_x_text() +
    labs(title = interacting_pair_str) +
    theme(legend.position = "bottom")
}
```

## Plot several interacting pairs

```{r, fig.asp = 1}
plot_interacting_pair(cpdb_tidy, "VCAM1_integrin_a4b1_complex", "Tumor")
plot_interacting_pair(cpdb_tidy, "CXCL8_CXCR2", "Neutrophil")
plot_interacting_pair(cpdb_tidy, "CXCL8_CXCR1", "Neutrophil")
plot_interacting_pair(cpdb_tidy, "CXCL8_ACKR1", "Neutrophil")
plot_interacting_pair(cpdb_tidy, "CD40_CD40LG", "DC")
plot_interacting_pair(cpdb_tidy, "CD40_CD40LG", "Mac.mono")
```

# Cell type specific, with stromal details

## Read in 

```{r}
cpdb_subdirs <- list.files(file.path(results_dir, "cpdb")) %>% 
  str_subset("cpdb") %>% 
  str_subset("10x_dir", negate = T)

cpdb_compartment_means <- cpdb_subdirs %>% 
  set_names() %>% 
  map_dfr(~read_tsv(file.path(results_dir, "cpdb", ..1, "out", "cell_type_specific_stromal_imm", "means.txt")), .id = "id")

cpdb_compartment_pvals <- cpdb_subdirs %>% 
  set_names() %>% 
  map_dfr(~read_tsv(file.path(results_dir, "cpdb", ..1, "out", "cell_type_specific_stromal_imm", "pvalues.txt")), .id = "id")

cpdb_means_tidy <- cpdb_compartment_means %>% 
  pivot_longer(cols = str_subset(colnames(.), "\\|"))
cpdb_pvals_tidy <- cpdb_compartment_pvals %>% 
  pivot_longer(cols = str_subset(colnames(.), "\\|"))
cpdb_tidy_stromal.imm <- cpdb_means_tidy %>%
  left_join(
    cpdb_pvals_tidy,
    by = join_by(
      id,
      id_cp_interaction,
      interacting_pair,
      partner_a,
      partner_b,
      gene_a,
      gene_b,
      secreted,
      receptor_a,
      receptor_b,
      annotation_strategy,
      is_integrin,
      name
    ), 
    suffix = c(".mean", ".pval")
  )
```

## Export/save

```{r}
saveRDS(cpdb_tidy_stromal.imm, file.path(results_dir, "cpdb_tidy_stromal.imm.rds"))
```

# Properties

```{r}
cpdb_tidy_stromal.imm$id_cp_interaction %>% unique() %>% length()
cpdb_tidy_stromal.imm$interacting_pair %>% unique() %>% length()
```

# Summarize total interactions

```{r}
cpdb_tidy_stromal.imm$id %>% unique() %>% 
  map(~summarize_total_ints(cpdb_tidy_stromal.imm, ..1))
```

## Plot several interacting pairs

```{r, fig.asp = 1}
plot_interacting_pair(cpdb_tidy_stromal.imm, "VCAM1_integrin_a4b1_complex", "Tumor")
plot_interacting_pair(cpdb_tidy_stromal.imm, "CXCL8_CXCR2", "Neutrophil")
plot_interacting_pair(cpdb_tidy_stromal.imm, "CD40_CD40LG", "DC")
plot_interacting_pair(cpdb_tidy_stromal.imm, "CD40_CD40LG", "Mac.mono")
```

## Examine top interactions

```{r}
cpdb_tidy_stromal.imm_sig_df <- cpdb_tidy_stromal.imm %>% 
  group_by(id) %>% 
  mutate(p_BH = p.adjust(value.pval, method = "BH")) %>% 
  ungroup() %>% 
  filter(p_BH < 0.05) %>% 
  arrange(desc(value.mean))
```

Number of Fibroblast+Endothelial Interactions

```{r}
n_fibroblast_ints <- cpdb_tidy_stromal.imm_sig_df %>% 
  filter(name %>% str_detect("Fibroblasts")) %>% 
  mutate(int_with = str_remove(name, "Fibroblasts")) %>% 
  mutate(int_with = str_remove(int_with, "\\|")) %>% 
  group_by(id, int_with) %>% 
  summarise(count = n()) 
n_endothelial_ints <- cpdb_tidy_stromal.imm_sig_df %>% 
  filter(name %>% str_detect("Endothelial")) %>% 
  mutate(int_with = str_remove(name, "Endothelial cells")) %>% 
  mutate(int_with = str_remove(int_with, "\\|")) %>% 
  group_by(id, int_with) %>% 
  summarise(count = n()) 
```

### plot against hypoxia

Shows that in more hypoxic samples, there are more stromal-tumor interactions (both fibroblasts and endothelial cells). 

```{r}
mean_hypoxia_df <- readRDS(file.path(data_dir, "scRNAseq", "mean_hypoxia_df.rds"))
```

```{r, fig.asp = 1, fig.width = 8}
n_fibroblast_ints %>% 
  mutate(sample_ID = str_remove(id, "cpdb_")) %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  ggplot() + 
  aes(mean_hypoxia, count.x) + 
  geom_point(size = 2) + 
  facet_wrap(~int_with, ncol = 4, scale = "free") + 
  stat_smooth(method = "lm")  + 
  stat_cor() + 
  labs(y = "Number of signif. interaction", 
       title = "With Fibroblasts")
n_endothelial_ints %>% 
  mutate(sample_ID = str_remove(id, "cpdb_")) %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  ggplot() + 
  aes(mean_hypoxia, count.x) + 
  geom_point(size = 2) + 
  facet_wrap(~int_with, ncol = 4, scale = "free") + 
  stat_smooth(method = "lm") + 
  stat_cor() + 
  labs(y = "Number of signif. interaction", 
       title = "With ECs")
```

### which interactions

#### Fibroblasts, with Tumor

```{r}
cpdb_tidy_stromal.imm_sig_df %>% 
  mutate(sample_ID = str_remove(id, "cpdb_")) %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  filter(name == "Fibroblasts|Tumor" | name == "Tumor|Fibroblasts") %>% 
  named_group_split(hypoxia_hi.v.lo) %>% 
  map(~{
    ..1 %>% 
      group_by(name, interacting_pair) %>% 
      summarise(n = n(), mean_mean_value = mean(value.mean)) %>% 
      filter(n > 5)
  })
```

#### Fibroblasts, with Myeloid

```{r}
cpdb_tidy_stromal.imm_sig_df %>% 
  mutate(sample_ID = str_remove(id, "cpdb_")) %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  filter(name == "Fibroblasts|DC" | name == "DC|Fibroblasts") %>% 
  named_group_split(hypoxia_hi.v.lo) %>% 
  map(~{
    ..1 %>% 
      group_by(name, interacting_pair) %>% 
      summarise(n = n(), mean_mean_value = mean(value.mean)) %>% 
      filter(n > 5)
  })
cpdb_tidy_stromal.imm_sig_df %>% 
  mutate(sample_ID = str_remove(id, "cpdb_")) %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  filter(name == "Fibroblasts|Mac/mono" | name == "Mac/mono|Fibroblasts") %>% 
  named_group_split(hypoxia_hi.v.lo) %>% 
  map(~{
    ..1 %>% 
      group_by(name, interacting_pair) %>% 
      summarise(n = n(), mean_mean_value = mean(value.mean)) %>% 
      filter(n > 5)
  })
```

#### Endothelial, with Tumor

```{r}
cpdb_tidy_stromal.imm_sig_df %>% 
  mutate(sample_ID = str_remove(id, "cpdb_")) %>% 
  left_join(mean_hypoxia_df, by = c("sample_ID")) %>% 
  filter(name == "Endothelial cells|Tumor" | name == "Tumor|Endothelial cells") %>% 
  named_group_split(hypoxia_hi.v.lo) %>% 
  map(~{
    ..1 %>% 
      group_by(name, interacting_pair) %>% 
      summarise(n = n(), mean_mean_value = mean(value.mean)) %>% 
      filter(n > 5)
  })
```

# Session info

```{r}
sessionInfo()
```

