---
title: "UCell signature scoring"
execute: 
  freeze: auto
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
source(here("lib", "single_cell", "helper_Seurat.R"))

# document specific packages: 
library(Seurat)
library(UCell)

file_cache_dir <- file.path(cache_dir, "sc-processing-ucell")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

# Read in data

## Seurat objects 

```{r}
so_stromal <- readRDS(file.path(cache_dir, "sc-processing-sctransform-stromal", "so_stromal.rds"))
so_tumor <- readRDS(file.path(cache_dir, "sc-processing-sctransform-tumor", "so_tumor.rds"))
so_imm <- readRDS(file.path(cache_dir, "sc-processing-sctransform-immune", "so_immune.rds"))
```

## Read in signature scores

### Bagaev signatures

```{r}
signature_df <- read_excel(
  file.path(data_dir, "signatures", "1-s2.0-S1535610821002221-mmc2.xlsx"), 
  sheet = "Gene Description", 
  skip = 1
) %>% mutate(Gene = str_replace(Gene, "-", "."))

signature_list <- signature_df %>% 
  named_group_split(`Gene signature`) %>% 
  map(~str_replace(.$Gene, "\\.", "-")) 
```

### Select Hallmarks signatures

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
select_gs <- c(
  "HALLMARK_INTERFERON_GAMMA_RESPONSE", 
  "HALLMARK_INTERFERON_ALHPA_RESPONSE", 
  "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", 
  "HALLMARK_ANGIOGENESIS", 
  "HALLMARK_HYPOXIA"
)
select_gs_vec <- gs_vec[select_gs]
```

# UCell signature scoring

## Stromal

```{r}
ucell_stromal <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_stromal_scores.rds"), 
  function() {
    ScoreSignatures_UCell(so_stromal@assays$SCT@data,
                          features = c(signature_list, select_gs_vec))
  }
)
```

Warning: The following genes were not found and will be imputed to exp=0:
* IFNA2,IFNB1,IL21,FCRL5,TNFRSF13B,CX3CR1,IL12B,CD244,KLRK1,NCR1,FCGR3B,MPO,PGLYRP1,IL22,IL13,IL4,IL5,CCR8,CCL1,MARCHF1,WARS1,CCN1,CCN2,BRS3,CCN5,LALBA

## Tumor

```{r}
ucell_tumor <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_tumor_scores.rds"), 
  function() {
    ScoreSignatures_UCell(so_tumor@assays$SCT@data,
                          features = c(signature_list, select_gs_vec))
  }
)
```

Warning: The following genes were not found and will be imputed to exp=0:
* IFNA2,IL21,CCR3,IL12B,XCR1,CCR8,CCL1,MARCHF1,WARS1,CCN1,CCN2,PRSS2,B4GALNT2,BRS3,CCN5

## Immune

```{r}
ucell_imm <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_imm_scores.rds"), 
  function() {
    ScoreSignatures_UCell(so_imm@assays$SCT@data,
                          features = c(signature_list, select_gs_vec))
  }
)
```

Warning: The following genes were not found and will be imputed to exp=0:
* IFNA2,NOS2,CA9,IL5,MARCHF1,WARS1,BDNF,CCN1,CCN2,FOXC2,SGCG,B4GALNT2,BRS3,CCN5,EDN2,IGFBP1,INHA,LALBA,NCAN

## Aggregate

```{r}
ucell_scores_df <- bind_rows(
  ucell_imm %>% as_tibble(rownames = "cell"), 
  ucell_tumor %>% as_tibble(rownames = "cell"), 
  ucell_stromal %>% as_tibble(rownames = "cell")
)
ucell_path <- file.path(results_dir, "adata_ucell_scores_df.rds")
if (!file.exists(ucell_path)) saveRDS(ucell_scores_df, ucell_path)
```

# Additional signatures

## Benci et al. 

```{r}
ifng_gs <- read_csv(file.path(data_dir, "signatures", "ifng_gs.csv"))$IFNG.GS
isg_rs <- read_csv(file.path(data_dir, "signatures", "isg_rs.csv"))$ISG.RS

benci_list <- list(
  ISG.RS = isg_rs, 
  IFNG.GS = ifng_gs
)

```

## Stromal

```{r}
ucell_stromal_benci <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_stromal_benci.rds"), 
  function() {
    ScoreSignatures_UCell(so_stromal@assays$SCT@data,
                          features = c(benci_list))
  }
)
```

Warning: The following genes were not found and will be
                        imputed to exp=0:
* FTSJD2,KLRK1,PRIC285,MARCHF1,WARS1,CCN1,CCN2,PGLYRP1,BRS3,CCN5,LALBA

## Tumor

```{r}
ucell_tumor_benci <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_tumor_benci.rds"), 
  function() {
    ScoreSignatures_UCell(so_tumor@assays$SCT@data,
                          features = c(benci_list))
  }
)
```

Warning: The following genes were not found and will be
                        imputed to exp=0:
* FTSJD2,PRIC285,MARCHF1,WARS1,CCN1,CCN2,PRSS2,B4GALNT2,BRS3,CCN5

## Immune

```{r}
ucell_imm_benci <- run_or_readRDS(
  file.path(file_cache_dir, "ucell_imm_benci.rds"), 
  function() {
    ScoreSignatures_UCell(so_imm@assays$SCT@data,
                          features = c(benci_list))
  }
)
```

Warning: The following genes were not found and will be
                        imputed to exp=0:
* FTSJD2,PRIC285,MARCHF1,WARS1,BDNF,CCN1,CCN2,FOXC2,SGCG,B4GALNT2,BRS3,CCN5,EDN2,IGFBP1,INHA,LALBA,NCAN

## Aggregate

```{r}
ucell_benci_df <- bind_rows(
  ucell_imm_benci %>% as_tibble(rownames = "cell"), 
  ucell_tumor_benci %>% as_tibble(rownames = "cell"), 
  ucell_stromal_benci %>% as_tibble(rownames = "cell")
)
ucell_benci_path <- file.path(results_dir, "adata_ucell_benci_df.rds")
if (!file.exists(ucell_benci_path)) saveRDS(ucell_benci_df, ucell_benci_path)
```

