---
title: "Stratification by cell type signatures"
execute: 
  freeze: auto
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
source(here("lib", "helper.R"))

# document specific packages: 
library(GSVA)
library(liger)

file_cache_dir <- file.path(cache_dir, "bulk-analysis-immune-high-vs-low")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

```{r}
bulk_clin_metadata <- read_tsv(file.path(data_dir, "bulk", "bulk-clin-metadata.tsv")) %>% 
  drop_na(sample_RNA) %>% 
  filter(is.na(excluded_clin) & is.na(excluded_rna)) %>% 
  mutate(prior_BRAF = case_when(
    str_detect(prior_therapy, "BRAF|dabrafenib|Dabra|Braf") ~ "BRAF", 
    cohort == "2019 Gide" & sample_RNA == "ipiPD1_14_PRE_R1" ~ "BRAF", 
    cohort == "2019 Gide" & sample_RNA == "ipiPD1_17_PRE_R1" ~ "BRAF", 
    prior_therapy == "unknown" ~ "unknown", 
    TRUE ~ "no BRAF"
  ))

tpm_df <- readRDS(file.path(results_dir, "tpm_df_combatseq.rds")) %>% 
  filter(sample_RNA %in% bulk_clin_metadata$sample_RNA)
tpm_tidy <- tpm_df %>% 
  pivot_longer(cols = 2:ncol(.), 
               names_to = "gene", 
               values_to = "value")
tpm_matrix <- tpm_df %>% 
  column_to_rownames("sample_RNA") %>% 
  as.matrix() %>% 
  t() 

signature_df <- read_excel(
  file.path(data_dir, "signatures", "1-s2.0-S1535610821002221-mmc2.xlsx"), 
  sheet = "Gene Description", 
  skip = 1
) %>% mutate(Gene = str_replace(Gene, "-", "."))

signature_list <- signature_df %>% 
  named_group_split(`Gene signature`) %>% 
  map(~.$Gene) 
```

```{r}
bagaev_ssgsea_df <- run_or_readRDS(
  file.path(file_cache_dir, "bagaev_ssgsea.rds"), 
  function() {
    gsva(
      tpm_matrix,
      gset.idx.list = signature_list,
      method = "ssgsea",
      kcdf = "Gaussian"
    ) %>%
      as_tibble(rownames = "gs_name")
  }
)
bagaev_ssgsea_tidy <- bagaev_ssgsea_df %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_RNA", values_to = "score") %>% 
  dplyr::rename(NAME = gs_name) %>% 
  group_by(NAME) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup()
write_tsv(bagaev_ssgsea_df, file.path(results_dir, "bagaev_ssgsea_df.tsv"))
```

```{r}
h_mtx <- bagaev_ssgsea_tidy %>% 
  pivot_wider(id_cols = NAME, 
              names_from = sample_RNA, 
              values_from = zscore) %>% 
  column_to_rownames("NAME") %>% 
  as.matrix()

Heatmap(h_mtx, 
        show_column_names = FALSE, 
        column_split = 5)
```

```{r, fig.asp = 0.9}
#| fig-asp: 0.9

Heatmap(cor(t(h_mtx), method = "spearman"), 
        name = "spearman\ncorr. ", 
        col = colorRamp2(breaks = c(-1, 0, 1), 
                         colors = palette_cor), 
        row_split = 3, 
        column_split = 3)
```

```{r}
h_dist <- dist(h_mtx)
row_dend <- as.dendrogram(hclust(h_dist)) %>% 
  reorder(h_dist)
immune_signatures <- labels(row_dend[[2]][[2]])
immune_signatures

write_tsv(tibble(signatures = immune_signatures), 
          file.path(results_dir, "immune_signatures.tsv"))
```

```{r}
h_row_dist <- dist(h_mtx[immune_signatures,])
row_dend <- as.dendrogram(hclust(h_row_dist)) %>% 
  reorder(h_row_dist)
h_col_dist <- stats::dist(t(h_mtx[immune_signatures,]), method = "euclidean")
col_hclust <- hclust(h_col_dist)
col_dend <- as.dendrogram(col_hclust) %>% 
  reorder(h_col_dist, method = "OLO")

h <- Heatmap(
  h_mtx[immune_signatures, ],
  show_column_names = FALSE,
  cluster_rows = row_dend,
  column_dend_reorder = TRUE, 
  column_split = cutree(col_hclust, 2)
)

h_col_dend <- column_dend(draw(h))
```

```{r}
kcluster_df <- h_col_dend %>% 
  imap_dfr(~tibble(sample_RNA = labels(..1), 
                   kcluster = ..2)) %>% 
  mutate(immune_hi.v.lo = case_when(
    kcluster %in% c("1") ~ "immune high", 
    kcluster %in% c("2") ~ "immune low"
  ))
```

## Silhouette Score

```{r}
max_clusters <- 20
cluster_out <-
  map(set_names(2:max_clusters), ~ cutree(col_hclust, k = ..1))
sil_out <-
  map(cluster_out, ~ cluster::silhouette(..1, h_col_dist))
sil_width_df <- imap_dfr(sil_out, ~ data.frame(
  k = ..2,
  avg_sil_width = mean(as_tibble(..1)$sil_width)
))
sil_width_df %>%
  mutate(k = factor(k, levels = 2:max_clusters)) %>% 
  ggplot() +
  aes(k, avg_sil_width) +
  geom_col() +
  labs(x = "K",
       y = "Average Silhouette Width") +
  geom_hline(yintercept = 0.4, color = "red") + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.45)) + 
  theme(axis.line.x = element_blank(), 
        axis.ticks.x = element_blank())
```

## Calculate immune score

```{r}
immune_zscore_df <- bagaev_ssgsea_tidy %>% 
  filter(NAME %in% immune_signatures) %>% 
  group_by(sample_RNA) %>% 
  summarise(immune_zscore = mean(zscore)) %>% 
  mutate(immune_hi.v.lo_zscore = ifelse(immune_zscore > median(immune_zscore), 
                                        "immune high", 
                                        "immune low")) %>% 
  mutate(immune_hi.med.lo_zscore = case_when(
    immune_zscore > quantile(immune_zscore, 2/3) ~ "immune high", 
    immune_zscore >= quantile(immune_zscore, 1/3) ~ "immune med", 
    immune_zscore < quantile(immune_zscore, 1/3) ~ "immune low", 
    TRUE ~ NA_character_
  ))
immune_hi.v.lo_df <- kcluster_df %>% 
  left_join(immune_zscore_df, by = "sample_RNA")
saveRDS(immune_hi.v.lo_df, file.path(results_dir, "immune_hi.v.lo.rds"))
```

```{r}
bulk_clin_metadata <- bulk_clin_metadata %>%
  select(-immune_hi.v.lo, -immune_zscore, -immune_hi.v.lo_zscore) %>%
  left_join(immune_hi.v.lo_df, by = "sample_RNA")
```

# Immune score vs response

```{r}
response_imm_count_df <- bulk_clin_metadata %>%
  group_by(response_splitSD, immune_hi.v.lo, treatment) %>%
  count() %>%
  pivot_wider(
    id_cols = c(treatment, immune_hi.v.lo),
    names_from = response_splitSD,
    values_from = n, 
    values_fill = 0
  ) 
response_imm_count_df
```

```{r}
prop_df <- bulk_clin_metadata %>%
  drop_na(response_splitSD) %>%
  group_by(immune_hi.v.lo, treatment, response_splitSD) %>%
  count() %>%
  group_by(immune_hi.v.lo, treatment) %>%
  mutate(total = sum(n), prop = n / total) %>%
  # add_factors() %>%
  mutate(
    treatment = factor(
      treatment %>%
        paste0("\n(n=", total, ")") %>%
        str_wrap(width = 8),
      levels = names(palette_treatment) %>%
        paste0("\n(n=", total, ")") %>%
        str_wrap(width = 8)
    )
  )
prop_df %>%
  ggplot() +
  aes(treatment, prop) +
  geom_col(aes(fill = response_splitSD), width = 0.6) +
  facet_grid( ~ immune_hi.v.lo, scale = "free_x") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  scale_fill_manual(values = palette_response) +
  labs(x = "Treatment", y = "Frequency") +
  coord_cartesian(clip = "off") +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "top",
    strip.background = element_blank()
  )
```

```{r}
calculate_fisher_response_by_treatment <- function(clin_metadata) {
  fisher_df <- clin_metadata %>%
    drop_na(response_splitSD) %>%
    group_by(treatment, response_splitSD, immune_hi.v.lo) %>%
    count() %>%
    drop_na(response_splitSD) %>%
    pivot_wider(
      id_cols = c("treatment", "immune_hi.v.lo"),
      names_from = response_splitSD,
      values_from = n
    ) %>%
    named_group_split(treatment) %>%
    map_dfr(
      ~ ..1 %>%
        dplyr::select(-treatment) %>%
        column_to_rownames("immune_hi.v.lo") %>%
        .[c("immune high", "immune low"), c("R", "NR")] %>%
        fisher.test() %>%
        broom::tidy(),
      .id = "treatment"
    ) %>%
    mutate(p.adj = p.adjust(p.value, method = "BH"))
  return(fisher_df)
}
fisher_df <- calculate_fisher_response_by_treatment(bulk_clin_metadata)
```

```{r}
fisher_df %>% 
  mutate(treatment = str_wrap(treatment, width = 12)) %>%
  mutate(p.value.symbol = case_when(
    p.value < 0.001 ~ "***", 
    p.value < 0.01 ~ "**",  
    p.value < 0.05 ~ "*", 
    TRUE ~ "n.s."
  )) %>% 
  ggplot() + 
  aes(treatment, estimate) + 
  geom_col(width = 0.5, size = 0.2) +
  geom_text(aes(label = format(estimate, digits = 3), y = 0.05),
            color = "white",
            hjust = 0) +
  geom_text(aes(label = sprintf("%s (p=%.3f)", p.value.symbol, p.value)),
            color = "black",
            hjust = -0.1) +
  geom_hline(yintercept = 1,
             color = "grey",
             linetype = "dashed") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 14)) +
  labs(x = "Treatment", y = "Odds Ratio (Immune high / immune low)") +
  coord_flip() +
  theme(axis.ticks.y = element_blank(),
        axis.line.y = element_blank())
```

# DGE

```{r}
bulk_clin_metadata_dge <- bulk_clin_metadata %>%
  mutate(responder = case_when(
    response == "R" ~ 1, 
    response == "NR" ~ 0, 
    TRUE ~ NA_integer_
  )) %>% 
  drop_na(responder) %>% 
  dplyr::select(sample_RNA, responder, treatment, immune_hi.v.lo)

bulk_clin_metadata_dge %>%
  group_by(responder, treatment, immune_hi.v.lo) %>% 
  count() %>% 
  pivot_wider(id_cols = c(immune_hi.v.lo, treatment), 
              names_from = responder, 
              values_from = n, 
              values_fill = 0) %>% 
  arrange(immune_hi.v.lo)
```

### Run DGE with custom functions

```{r}
dge_treatment_immune_df <- run_or_readRDS(
  file.path(file_cache_dir, "dge_R.v.NR_by_treatment_immune.rds"), 
  function() {
    run_dge_split(
      tpm_tidy,
      bulk_clin_metadata_dge %>%
        unite("treatment_immune", treatment, immune_hi.v.lo) %>%
        filter(treatment_immune != "combo_immune high"),
      responder,
      sample_RNA,
      treatment_immune
    ) %>%
      separate(treatment_immune,
               into = c("treatment", "immune_hi.v.lo"),
               sep = "_")
  }
)
dge_treatment_immune_df %>% 
  write_tsv(file.path(results_dir, "dge_treatment_immune_df.tsv"))
```


## No significant results

At a threshold of FDR less than 0.05, there are no significant genes across all groups. 
```{r}
dge_treatment_immune_df %>% 
  filter(p_BH < 0.05)
```

If we apply a more generous filter (FDR less then 0.25), we are still unable to find many significantly differentially expressed genes. 

```{r}
dge_treatment_immune_df %>% 
  filter(p_BH < 0.25)
```

# Gene Set Enrichment Analysis

## Read in gene sets

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
```

## Run GSEA using liger

```{r}
gsea_treatment_immune_df <- run_or_readRDS(
  file.path(file_cache_dir, "gsea_R.v.NR_by_treatment_immune.rds"), 
  function() {
    dge_list <- dge_treatment_immune_df %>% 
      arrange(desc(signed_neglog10_p)) %>% 
      named_group_split(treatment, immune_hi.v.lo) %>% 
      map(~set_names(.$signed_neglog10_p, .$gene))
    gsea_list <- dge_list %>% 
      map(iterative.bulk.gsea, set.list = gs_vec, power = 0)
    gsea_df <- gsea_list %>% 
      map_dfr(as_tibble, rownames = "gs_name", .id = "treatment_immune") %>% 
      separate("treatment_immune", into = c("treatment", "immune_hi.v.lo"), sep = " / ")
  }
)
write_tsv(gsea_treatment_immune_df, 
          file.path(results_dir, "gsea_R.v.NR_by_treatment_immune.tsv"))
```

## GSEA results

### What signatures

```{r}
gsea_treatment_immune_df %>% 
  named_group_split(treatment, immune_hi.v.lo) %>% 
  imap(~{
    ..1 %>%
      mutate(direction = ifelse(sscore > 0, "R", "NR")) %>%
      ggplot() +
      aes(fct_reorder(gs_name, sscore), sscore, color = q.val < 0.05) +
      geom_point(shape = "|", size = 3) +
      geom_text_repel(aes(label = gs_name),
                data = ~ ..1 %>%
                  filter(q.val < 0.05) %>% 
                  filter(direction == "R") %>% 
                  slice_max(abs(sscore), n = 4), 
                hjust = 1) +
      geom_text_repel(aes(label = gs_name),
                data = ~ ..1 %>%
                  filter(q.val < 0.05) %>% 
                  filter(direction == "NR") %>% 
                  slice_max(abs(sscore), n = 4), 
                hjust = 1) +
      scale_color_manual(values = palette_bool) +
      remove_x_spine() +
      theme(axis.text.x = element_blank()) + 
      labs(title = ..2)
  })
```

```{r, fig.asp = 1}
#| fig-asp: 1

gsea_treatment_immune_wide <- gsea_treatment_immune_df %>% 
  pivot_wider(id_cols = c(gs_name, immune_hi.v.lo), 
              names_from = treatment, 
              values_from = c(q.val, sscore)) 
p <- gsea_treatment_immune_wide %>% 
  filter(immune_hi.v.lo == "immune high") %>% 
  ggplot() + 
  aes(sscore_aPD1, sscore_aCTLA4, shape = q.val_aPD1 < 0.1 & q.val_aCTLA4 < 0.1, label = gs_name) + 
  geom_point(size = 2) + 
  scale_shape_manual(values = c(4, 19)) + 
  theme(legend.position = "top")
p + 
  geom_point(size = 3, data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA"), color = "red") + 
  geom_text_repel(data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA"), color = "red") 
ggplotly(p + theme(legend.position = "none"))
```

```{r, fig.asp = 1}
#| fig-asp: 1

gsea_treatment_immune_wide <- gsea_treatment_immune_df %>% 
  pivot_wider(id_cols = c(gs_name, treatment), 
              names_from = immune_hi.v.lo, 
              values_from = c(q.val, sscore)) 
p <- gsea_treatment_immune_wide %>% 
  filter(treatment == "aPD1") %>% 
  ggplot() + 
  aes(`sscore_immune low`, `sscore_immune high`, label = gs_name, 
      shape = `q.val_immune low` < 0.1 & `q.val_immune high` < 0.1) + 
  geom_point(size = 2) + 
  geom_point(color = "red", 
             data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA")) + 
  scale_shape_manual(values = c(4, 19)) + 
  theme(legend.position = "top")
p +
  geom_point(size = 3, data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA"), color = "red") + 
  geom_text_repel(data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA"), color = "red") 
ggplotly(p + theme(legend.position = "none"))
```

## Change from previous

```{r}
gsea_treatment_df <- readRDS(file.path(cache_dir, "bulk-analysis-dge-and-gsea", "gsea_R.v.NR_by_treatment.rds"))
```

```{r, fig.asp = 1}
plot_df <- gsea_treatment_immune_df %>% 
  bind_rows(gsea_treatment_df) %>% 
  mutate(signed_p = ifelse(sscore > 0, -log10(p.val), log10(p.val))) %>% 
  group_by(treatment, immune_hi.v.lo) %>% 
  arrange(desc(signed_p), desc(sscore)) %>% 
  mutate(rank = row_number()) %>% 
  pivot_wider(id_cols = c(treatment, gs_name), 
              names_from = immune_hi.v.lo, 
              values_from = c(q.val, rank, sscore)) %>% 
  filter(treatment == "aPD1") %>% 
  mutate(direction_pre = ifelse(sscore_NA > 0, "enriched in R", "enriched in NR")) %>% 
  mutate(immhigh_delta = `rank_NA` - `rank_immune high`) %>% 
  mutate(immlow_delta = `rank_NA` - `rank_immune low`)
# plot_df
plot_delta_lollipop <- function(plot_df, var_str) {
  plot_df %>% 
    filter(abs(get(var_str)) < 100) %>% 
    ggplot() + 
    aes(fct_reorder(gs_name, get(var_str)), get(var_str), 
        color = direction_pre) + 
    geom_hline(yintercept = 0, linetype = "dashed") + 
    geom_segment(aes(xend = gs_name, y = 0, yend = get(var_str))) + 
    geom_point(size = 3) + 
    turn_xaxis_labels() + 
    remove_x_spine() + 
    theme(panel.grid.major.x = element_line()) + 
    labs(y = var_str, title = var_str, subtitle = "signif in before/after") 
}
plot_delta_lollipop(plot_df %>%
                      filter(`q.val_immune high` < 0.05 &
                               `q.val_NA` < 0.05),
                    "immhigh_delta")
plot_delta_lollipop(plot_df %>%
                      filter(`q.val_immune low` < 0.05 &
                               `q.val_NA` < 0.05),
                    "immlow_delta") + 
  theme(axis.text.x = element_text(size = 5))
```

```{r}
gsea_treatment_immune_df %>% 
  filter(gs_name == "HALLMARK_HYPOXIA")
```

```{r}
gsea_treatment_immune_df %>% 
  filter(gs_name == "HALLMARK_INTERFERON_GAMMA_RESPONSE")
```

