---
title: "Hypoxia"
execute: 
  freeze: auto
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))

# document specific packages: 
library(GSVA)

file_cache_dir <- file.path(cache_dir, "bulk-analysis-hypoxia")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

# Read in data

First, we have to read in the data tables which have been generated by previous scripts. 

```{r}
immune_hi.v.lo_df <- readRDS(file.path(results_dir, "immune_hi.v.lo.rds"))
bulk_clin_metadata <- read_tsv(file.path(data_dir, "bulk", "bulk-clin-metadata.tsv")) %>% 
  drop_na(sample_RNA) %>% 
  filter(is.na(excluded_clin) & is.na(excluded_rna)) %>% 
  select(-immune_hi.v.lo, -immune_zscore, -immune_hi.v.lo_zscore) %>% 
  left_join(immune_hi.v.lo_df, by = "sample_RNA")

dge_treatment <- read_tsv(file.path(results_dir, "dge_treatment.tsv"))
gsea_treatment <- read_tsv(file.path(results_dir, "gsea_treatment.tsv"))
gsea_treatment_imm <- read_tsv(file.path(results_dir, "gsea_R.v.NR_by_treatment_immune.tsv"))

ssgsea_tidy <- readRDS(file.path(results_dir, "ssgsea_tidy.rds"))
ssgsea_tidy_clin <- ssgsea_tidy %>% 
  left_join(bulk_clin_metadata, by = "sample_RNA") %>% 
  drop_na(patient, immune_hi.v.lo)
```

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
```

# Without immune stratification, hypoxia is enriched in NR to aPD1

We can look at where the distributions of hypoxia related genes are in the volcano plots. 

```{r}
dge_treatment %>% 
  mutate(is_in_hallmark_hypoxia = gene %in% gs_vec$HALLMARK_HYPOXIA) %>% 
  ggplot() + 
  aes(log2(med_fc), -log10(p.value)) +
  geom_scattermore() + 
  geom_point(color = "red", data = ~..1 %>% filter(is_in_hallmark_hypoxia)) + 
  geom_text_repel(aes(label = gene), 
                  data = ~..1 %>% filter(gene %in% c("HIF1A", "EPAS1"))) + 
  facet_wrap(~treatment, ncol = 2) + 
  scale_color_manual(values = palette_bool) + 
  theme(legend.position = "top")
```

```{r, fig.asp = 1, fig.width = 4}
#| fig-asp: 1
#| fig-width: 4

plot_gsea_lollipop_by_treatment <- function(gsea_df, gs_name_str) {
  gsea_df %>%
    filter(gs_name == gs_name_str) %>%
    ggplot() +
    aes(treatment, sscore) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_point(size = 3, aes(shape = q.val < 0.05)) +
    geom_segment(aes(
      y = 0,
      yend = sscore,
      x = treatment,
      xend = treatment
    )) +
    remove_x_spine() +
    turn_xaxis_labels() +
    scale_shape_manual(values = c(4, 19)) + 
    labs(title = gs_name_str)
}

plot_gsea_lollipop_by_treatment(gsea_treatment, "HALLMARK_HYPOXIA")
```

```{r}
ssgsea_tidy_clin %>% 
  filter(NAME == "HALLMARK_HYPOXIA") %>% 
  drop_na(response) %>% 
  ggplot() + 
  aes(response, zscore) + 
  geom_boxplot(outlier.size = 0, outlier.stroke = 0) + 
  geom_jitter(width = 0.2) + 
  facet_grid(~treatment) + 
  stat_compare_means(comparison = list(c("R", "NR"))) + 
  remove_x_spine()
```

Where is hypoxia in the context of all the other gene sets that we examined? Below is showing just aPD1. 

```{r}
gsea_treatment %>% 
  filter(treatment == "aPD1") %>% 
  filter(q.val < 0.05) %>% 
  ggplot() + 
  aes(fct_reorder(gs_name, sscore), sscore) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_segment(aes(xend = gs_name, y = 0, yend = sscore)) + 
  geom_point(size = 2) + 
  geom_point(color = "red", 
             data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA"), 
             size = 2.5) + 
  coord_flip() + 
  theme(axis.text.y = element_text(size = 5)) + 
  remove_spines() + 
  labs(x = "gene set")
```

# NR immune high patients MIGHT enrich in hypoxia more than NR immune low patients

From our GSEA calculations, we see that hypoxia enriched in immune high non-responders relative to responders, but NOT in immune low. 

```{r}
plot_gsea_lollipop_by_treatment(gsea_treatment_imm, "HALLMARK_HYPOXIA") + 
  aes(color = immune_hi.v.lo) + 
  facet_grid( ~ immune_hi.v.lo) + 
  scale_color_manual(values = palette_immune_state)
```

It's worth noting that in terms of significantly enriched pathways in immune high aPD1, hypoxia is NOT one of the top pathways for aPD1 though. 

```{r}
gsea_treatment_imm %>% 
  filter(treatment == "aPD1" & immune_hi.v.lo == "immune high") %>% 
  filter(q.val < 0.05) %>% 
  ggplot() + 
  aes(fct_reorder(gs_name, sscore), sscore) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_segment(aes(xend = gs_name, y = 0, yend = sscore)) + 
  geom_point(size = 2) + 
  geom_point(color = "red", 
             data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA"), 
             size = 2.5) + 
  coord_flip() + 
  theme(axis.text.y = element_text(size = 7)) + 
  remove_spines() + 
  labs(x = "gene set")
```


```{r}
gsea_treatment_imm %>% 
  filter(treatment == "aPD1" & immune_hi.v.lo == "immune low") %>% 
  filter(q.val < 0.05) %>% 
  ggplot() + 
  aes(fct_reorder(gs_name, sscore), sscore) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_segment(aes(xend = gs_name, y = 0, yend = sscore)) + 
  geom_point(size = 2) + 
  coord_flip() + 
  theme(axis.text.y = element_text(size = 5)) + 
  remove_spines() + 
  labs(x = "gene set")
```

This slight difference in immune high and low is apparent when looking at the ssGSEA scores, which consider signatures on a per-sample basis as well. Below is showing just aPD1. 

```{r}
plot_ssgsea_by_treatment_and_response <- function(gs_str) {
  ssgsea_tidy_clin %>%
    filter(NAME == gs_str) %>%
    drop_na(response) %>%
    ggplot() +
    aes(response, score) +
    geom_violin(scale = "width", draw_quantiles = 0.5) +
    geom_jitter(width = 0.2, size = 2, aes(color = response)) +
    facet_grid( ~ treatment) +
    theme(axis.text.x = element_text(
      angle = 90,
      hjust = 1,
      vjust = 0.5
    )) +
    scale_color_manual(values = palette_response) + 
    stat_compare_means() +
    remove_x_spine() + 
    labs(title = gs_str)
}
plot_ssgsea_by_treatment_and_response("HALLMARK_HYPOXIA") + 
  facet_grid(~immune_hi.v.lo) + 
  labs(subtitle = "stratified by hierarchical clustering")
```

```{r}
plot_ssgsea_by_treatment_and_response("HALLMARK_HYPOXIA") + 
  facet_grid(~immune_hi.v.lo_zscore) + 
  labs(subtitle = "stratified by median immune score")
plot_ssgsea_by_treatment_and_response("HALLMARK_HYPOXIA") + 
  facet_grid(~immune_hi.med.lo_zscore) + 
  labs(subtitle = "stratified by tertile immune score")
```

For sure, when split by the median or into terciles, the difference between how hypoxia affects response seems to differ more greatly between immune high and immune low. 

## Immune score as a continuous variables

```{r}
plot_df <- ssgsea_tidy_clin %>% 
  drop_na(response) %>% 
  filter(treatment == "aPD1") %>% 
  filter(NAME == "HALLMARK_HYPOXIA") 

# plot_df %>% 
#   group_by(response) %>% 
#   summarise(lm = list(summary(lm(zscore ~ immune_zscore)))) %>% 
#   .[["lm"]]

plot_df %>% 
  ggplot() + 
  aes(immune_zscore, zscore, color = response) + 
  geom_point(size = 3) + 
  stat_ellipse() + 
  stat_smooth(method = "lm") + 
  scale_color_manual(values = palette_response) + 
  labs(y = "HALLMARK HYPOXIA") + 
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 1.2, 
    vjust = 1.2, 
    label = c("y = 0.53 + 0.22x, p = 0.0012"), 
    color = palette_response[["NR"]]
  ) + 
  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 1.2, 
    vjust = 3.2, 
    label = "y = 0.15 - 0.21x, p = 0.273", 
    color = palette_response[["R"]]
  )


```

While non-responders and responders have both immune high and immune low samples, immune high non-responders will tend to have higher hypoxia too, as evidenced by the yellow fit line above.  

## Modeling suggests hypoxia affects response to aPD1 but does not confirm immune involvement

This could be aided by some type of modeling. 

We're going to just focus on aPD1 here. First, we should absolutely consider the immune score as a continuous variable: 

```{r}
models <- list(
  "response ~ 1", 
  "response ~ zscore", 
  "response ~ immune_zscore", 
  "response ~ zscore:immune_zscore", 
  "response ~ immune_zscore + zscore", 
  "response ~ immune_zscore * zscore"
) %>% set_names() 

ssgsea_tidy_treatment_list <- ssgsea_tidy_clin %>%
  filter(NAME %in% c("HALLMARK_HYPOXIA")) %>%
  drop_na(response) %>%
  mutate(response = as.integer(as.character(response) == "R")) %>%
  named_group_split(treatment)

model_list <- models %>%
  map(~ glm(
    formula(..1),
    family = "binomial",
    data = ssgsea_tidy_treatment_list$aPD1
  ))

model_list %>% 
  map_dfr(AIC, .id = "treatment") %>% 
  t() %>% 
  as_tibble(rownames = "model") %>% 
  dplyr::rename(AIC = V1) %>% 
  arrange(AIC) %>% 
  mutate(delta_AIC = AIC - AIC[1])
```

What if we consider immune state (discrete) instead of immune score (continuous):

```{r}
models <- list(
  "response ~ 1", 
  "response ~ zscore", 
  "response ~ immune_hi.v.lo_zscore", 
  "response ~ immune_hi.v.lo_zscore:zscore", 
  "response ~ immune_hi.v.lo_zscore + zscore", 
  "response ~ immune_hi.v.lo_zscore * zscore"
) %>% set_names() 

ssgsea_tidy_treatment_list <- ssgsea_tidy_clin %>%
  filter(NAME %in% c("HALLMARK_HYPOXIA")) %>%
  drop_na(response) %>%
  mutate(response = as.integer(as.character(response) == "R")) %>%
  mutate(immune_hi.med.lo_zscore = factor(
    immune_hi.med.lo_zscore,
    levels = c("immune low", "immune med", "immune high"),
    ordered = TRUE
  )) %>% 
  mutate(immune_hi.v.lo_zscore = factor(
    immune_hi.v.lo_zscore,
    levels = c("immune low", "immune high"),
    ordered = TRUE
  )) %>% 
  named_group_split(treatment)

model_list <- models %>%
  map(~ glm(
    formula(..1),
    family = "binomial",
    data = ssgsea_tidy_treatment_list$aPD1
  ))

model_list %>% map_dfr(broom::tidy, .id = "model")

model_list %>% 
  map_dfr(AIC, .id = "treatment") %>% 
  t() %>% 
  as_tibble(rownames = "model") %>% 
  dplyr::rename(AIC = V1) %>% 
  arrange(AIC) %>% 
  mutate(delta_AIC = AIC - AIC[1])
```

Just the hypoxia score is the best model. Makes sense!

```{r}
models <- list(
  "response ~ 1", 
  "response ~ zscore"
) %>% set_names() 

ssgsea_tidy_treatment_list <- ssgsea_tidy_clin %>%
  filter(NAME %in% c("HALLMARK_HYPOXIA")) %>%
  drop_na(response) %>%
  mutate(response = as.integer(as.character(response) == "R")) %>%
  mutate(immune_hi.v.lo_zscore = factor(
    immune_hi.v.lo_zscore,
    levels = c("immune low", "immune high"),
    ordered = TRUE
  )) %>% 
  named_group_split(treatment, immune_hi.v.lo)

model_list <- c(
  models %>%
    map(
      ~ glm(
        formula(..1),
        family = "binomial",
        data = ssgsea_tidy_treatment_list$`aPD1 / immune low`
      )
    ) %>% set_names(paste0("immune low, ", models)),
  models %>%
    map(
      ~ glm(
        formula(..1),
        family = "binomial",
        data = ssgsea_tidy_treatment_list$`aPD1 / immune high`
      )
    ) %>% set_names(paste0("immune high, ", models))
)

model_list %>% map_dfr(broom::tidy, .id = "model")
```

