---
title: "DGE + GSEA"
execute: 
  freeze: auto
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))
source(here("lib", "helper.R"))

# document specific packages: 
library(liger)

file_cache_dir <- file.path(cache_dir, "bulk-analysis-dge-and-gsea")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

```{r}
bulk_clin_metadata <- read_tsv(file.path(data_dir, "bulk", "bulk-clin-metadata.tsv")) %>% 
  drop_na(sample_RNA) %>% 
  filter(is.na(excluded_clin) & is.na(excluded_rna))
tpm_df <- readRDS(file.path(results_dir, "tpm_df_combatseq.rds"))
tpm_tidy <- tpm_df %>% 
  pivot_longer(cols = 2:ncol(.), 
               names_to = "gene", 
               values_to = "value")
```

```{r}
bulk_clin_metadata_dge <- bulk_clin_metadata %>%
  mutate(responder = case_when(
    response == "R" ~ 1, 
    response == "NR" ~ 0, 
    TRUE ~ NA_integer_
  )) %>% 
  drop_na(responder) %>% 
  dplyr::select(sample_RNA, responder, treatment)

bulk_clin_metadata_dge %>%
  group_by(responder, treatment) %>% 
  count() %>% 
  pivot_wider(id_cols = c(responder), 
              names_from = treatment, 
              values_from = n, 
              values_fill = 0)
```

# Run DGE by treatment

## Run DGE with custom functions

```{r}
dge_treatment_df <- run_or_readRDS(
  file.path(file_cache_dir, "dge_R.v.NR_by_treatment.rds"), 
  function() {
    run_dge_split(
      tpm_tidy,
      bulk_clin_metadata_dge,
      responder,
      sample_RNA,
      treatment
    )
  }
)
```

## No significant results

At a threshold of FDR less than 0.05, there are no significant genes across all groups. 

```{r}
dge_treatment_df %>% 
  filter(p_BH < 0.05)
```

If we apply a more generous filter (FDR less then 0.25), we are still unable to find significantly differentially expressed genes. 

```{r}
dge_treatment_df %>% 
  filter(p_BH < 0.25)
```

# Gene Set Enrichment Analysis

## Read in gene sets

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
```

## Run GSEA using liger

```{r}
gsea_treatment_df <- run_or_readRDS(
  file.path(file_cache_dir, "gsea_R.v.NR_by_treatment.rds"), 
  function() {
    dge_list <- dge_treatment_df %>% 
      arrange(desc(signed_neglog10_p)) %>% 
      named_group_split(treatment) %>% 
      map(~set_names(.$signed_neglog10_p, .$gene))
    gsea_list <- dge_list %>% 
      map(iterative.bulk.gsea, set.list = gs_vec, power = 0)
    gsea_df <- gsea_list %>% 
      map_dfr(as_tibble, rownames = "gs_name", .id = "treatment")
  }
)
```

```{r}
gsea_treatment_df %>% 
  filter(q.val < 0.05) %>% 
  mutate(direction = ifelse(sscore > 0, "R", "NR")) %>% 
  group_by(treatment, direction) %>% 
  count() %>% 
  pivot_wider(id_cols = c(treatment), 
              names_from = direction, 
              values_from = n) %>% 
  mutate(total_signif = NR + R)
```

## GSEA results

```{r}
gsea_treatment_df %>% 
  named_group_split(treatment) %>% 
  imap(~{
    ..1 %>%
      mutate(direction = ifelse(sscore > 0, "R", "NR")) %>%
      ggplot() +
      aes(fct_reorder(gs_name, sscore), sscore, color = q.val < 0.05) +
      geom_point(shape = "|", size = 3) +
      geom_text_repel(aes(label = gs_name),
                data = ~ ..1 %>%
                  filter(q.val < 0.05) %>% 
                  filter(direction == "R") %>% 
                  slice_max(abs(sscore), n = 4), 
                hjust = 1) +
      geom_text_repel(aes(label = gs_name),
                data = ~ ..1 %>%
                  filter(q.val < 0.05) %>% 
                  filter(direction == "NR") %>% 
                  slice_max(abs(sscore), n = 4), 
                hjust = 1) +
      scale_color_manual(values = palette_bool) +
      remove_x_spine() +
      theme(axis.text.x = element_blank()) + 
      labs(title = ..2)
  })
```

```{r}
gsea_treatment_wide <- gsea_treatment_df %>% 
  pivot_wider(id_cols = gs_name, 
              names_from = treatment, 
              values_from = c(sscore, q.val))
```

```{r}
p <- gsea_treatment_wide %>% 
  ggplot() + 
  aes(sscore_aCTLA4, sscore_combo, 
      label = gs_name, 
      shape = q.val_aCTLA4 < 0.05 & q.val_combo < 0.05) + 
  geom_point(size = 3) + 
  theme(legend.position = "none") + 
  scale_shape_manual(values = c(4, 19)) +
  geom_point(size = 3, color = "red",
             data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA"))
ggplotly(p)
```

```{r}
p <- gsea_treatment_wide %>% 
  ggplot() + 
  aes(sscore_aPD1, sscore_combo, 
      label = gs_name, 
      shape = q.val_aPD1 < 0.05 & q.val_combo < 0.05) + 
  geom_point(size = 3) + 
  theme(legend.position = "none") + 
  scale_shape_manual(values = c(4, 19)) +
  geom_point(size = 3, color = "red",
             data = ~..1 %>% filter(gs_name == "HALLMARK_HYPOXIA"))
ggplotly(p)
```

# Save

```{r}
if (!file.exists(file.path(results_dir, "dge_treatment.tsv"))) {
  write_tsv(
    dge_treatment_df,
    file.path(results_dir, "dge_treatment.tsv")
  )
}
if (!file.exists(file.path(results_dir, "gsea_treatment.tsv"))) {
  write_tsv(
    gsea_treatment_df,
    file.path(results_dir,
              "gsea_treatment.tsv")
  )
}
```

