---
title: "ssGSEA analysis"
execute: 
  freeze: true 
  message: false
  warning: false
---

```{r}
library(here)
source(here("lib", "setup.R"))
source(here("lib", "aesthetics.R"))

# document specific packages: 
library(GSVA)

file_cache_dir <- file.path(cache_dir, "bulk-analysis-ssgsea")
if (!dir.exists(file_cache_dir)) dir.create(file_cache_dir)
```

# Read in data

```{r}
immune_hi.v.lo_df <- readRDS(file.path(results_dir, "immune_hi.v.lo.rds"))
bulk_clin_metadata <- read_tsv(file.path(data_dir, "bulk", "bulk-clin-metadata.tsv")) %>% 
  drop_na(sample_RNA) %>% 
  filter(is.na(excluded_clin) & is.na(excluded_rna)) %>% 
  select(-immune_hi.v.lo, -immune_zscore, -immune_hi.v.lo_zscore) %>% 
  left_join(immune_hi.v.lo_df, by = "sample_RNA")
tpm_df <- readRDS(file.path(results_dir, "tpm_df_combatseq.rds")) %>% 
  filter(sample_RNA %in% bulk_clin_metadata$sample_RNA)
tpm_matrix <- tpm_df %>% 
  column_to_rownames("sample_RNA") %>% 
  as.matrix() %>% 
  t()
```

```{r}
gs_vec <- readRDS(file.path(results_dir, "gs_vec.rds"))
hypoxia_vec <- readRDS(file.path(results_dir, "hypoxia_gs_vec.rds"))
```

# Run ssGSEA

## On broad set of signatures

```{r}
ssgsea_df <- run_or_readRDS(
  file.path(file_cache_dir, "ssgsea.rds"), 
  function() {
    gsva(
      tpm_matrix,
      gset.idx.list = gs_vec,
      method = "ssgsea",
      kcdf = "Gaussian",
      min.sz = 3
    ) %>%
      as_tibble(rownames = "gs_name")
  }
)
ssgsea_tidy <- ssgsea_df %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_RNA", values_to = "score") %>% 
  dplyr::rename(NAME = gs_name) %>% 
  group_by(NAME) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup()
write_tsv(ssgsea_df, file.path(results_dir, "ssgsea_df.tsv"))
saveRDS(ssgsea_tidy, file.path(results_dir, "ssgsea_tidy.rds"))
```

# On extra hypoxia signatures

```{r}
ssgsea_hypoxia_df <- run_or_readRDS(
  file.path(file_cache_dir, "ssgsea_hypoxia.rds"), 
  function() {
    gsva(
      tpm_matrix,
      gset.idx.list = hypoxia_vec,
      method = "ssgsea",
      kcdf = "Gaussian",
      min.sz = 3
    ) %>%
      as_tibble(rownames = "gs_name")
  }
)
ssgsea_hypoxia_tidy <- ssgsea_hypoxia_df %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_RNA", values_to = "score") %>% 
  dplyr::rename(NAME = gs_name) %>% 
  group_by(NAME) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup()
write_tsv(ssgsea_hypoxia_df, file.path(results_dir, "ssgsea_hypoxia_df.tsv"))
saveRDS(ssgsea_hypoxia_tidy, file.path(results_dir, "ssgsea_hypoxia_tidy.rds"))
```

# View all results

```{r}
h_mtx <- ssgsea_tidy %>% 
  pivot_wider(id_cols = NAME, 
              names_from = sample_RNA, 
              values_from = zscore) %>% 
  column_to_rownames("NAME") %>% 
  as.matrix()
```

```{r, fig.asp = 1/2, fig.width = 20}
Heatmap(
  h_mtx, 
  name = "scaled score", 
  show_row_names = TRUE, 
  show_column_names = FALSE, 
  row_names_gp = gpar(fontsize = 5)
)
```

```{r, fig.width = 12}
Heatmap(
  cor(t(h_mtx), method = "pearson"), 
  name = "pearson cor.", 
  show_column_names = FALSE, 
  row_names_gp = gpar(fontsize = 3), 
  height = unit(6, "in"), 
  width = unit(6, "in")
)
```

# VS Response?

```{r}
ssgsea_tidy_clin <- ssgsea_tidy %>% 
  left_join(bulk_clin_metadata, by = "sample_RNA") %>% 
  drop_na(patient, immune_hi.v.lo)
```

```{r}
ssgsea_tidy_clin %>% 
  drop_na(response) %>% 
  filter(NAME == "HALLMARK_INTERFERON_GAMMA_RESPONSE") %>% 
  ggplot() + 
  aes(response, score) + 
  geom_boxplot(outlier.size = 0, outlier.stroke = 0) + 
  geom_jitter(width = 0.2) + 
  stat_compare_means(comparison = list(c("R", "NR"))) + 
  scale_y_continuous(expand = c(0, 0), limits = ~c(..1[1], ..1[2]*1.2)) + 
  facet_grid(immune_hi.v.lo ~ treatment)
```

# Sanity checks

## Hallmark IFNg response == More IFNg expression

It's not trivial to detect IFNg transcriptionally (most cytokines are very transient), but, reassuringly, the expression of this signatures correlates very nicely with IFNG expression. 

```{r}
ifng_df <- tpm_df %>% 
  select(sample_RNA, IFNG)
ssgsea_tidy_clin %>% 
  filter(NAME == "HALLMARK_INTERFERON_GAMMA_RESPONSE") %>% 
  left_join(ifng_df, by = "sample_RNA") %>% 
  ggplot() + 
  aes(score, IFNG, color = immune_hi.v.lo) + 
  geom_point(size = 2, alpha = 0.5) + 
  stat_cor(method = "spearman") + 
  scale_color_manual(values = palette_immune_state)
```

# plot by treatment and response 

```{r}
plot_ssgsea_by_treatment_and_response <- function(ssgsea_tidy, gs_str) {
  ssgsea_tidy %>%
    filter(NAME == gs_str) %>%
    drop_na(response) %>%
    ggplot() +
    aes(response, score) +
    geom_violin(scale = "width", draw_quantiles = 0.5) +
    geom_jitter(width = 0.2) +
    facet_grid( ~ treatment) +
    theme(axis.text.x = element_text(
      angle = 90,
      hjust = 1,
      vjust = 0.5
    )) +
    stat_compare_means(comparisons = list(c("NR", "R"))) +
    remove_x_spine() + 
    labs(title = gs_str)
}
plot_ssgsea_by_treatment_and_response(ssgsea_tidy_clin, "HALLMARK_HYPOXIA")
plot_ssgsea_by_treatment_and_response(ssgsea_tidy_clin, "HALLMARK_ANGIOGENESIS")
plot_ssgsea_by_treatment_and_response(ssgsea_tidy_clin, "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION")
plot_ssgsea_by_treatment_and_response(ssgsea_tidy_clin, "HALLMARK_INTERFERON_GAMMA_RESPONSE")
```

# Extra hypoxia signatures

```{r}
ssgsea_hypoxia_tidy_clin <- 
  ssgsea_hypoxia_tidy %>% 
  left_join(bulk_clin_metadata, by = "sample_RNA") 

ssgsea_hypoxia_tidy_clin %>% 
  plot_ssgsea_by_treatment_and_response("BUFFA_HYPOXIA_METAGENE") + 
  facet_wrap( ~ immune_hi.v.lo + treatment, ncol = 4, scale = "free_y") + 
  scale_y_continuous(expand = c(0.2, 0))

ssgsea_hypoxia_tidy_clin %>% 
  plot_ssgsea_by_treatment_and_response("HARRIS_HYPOXIA") + 
  facet_wrap( ~ immune_hi.v.lo + treatment, ncol = 4, scale = "free_y") + 
  scale_y_continuous(expand = c(0.2, 0))

ssgsea_hypoxia_tidy_clin %>% 
  plot_ssgsea_by_treatment_and_response("WINTER_HYPOXIA_METAGENE") + 
  facet_wrap( ~ immune_hi.v.lo + treatment, ncol = 4, scale = "free_y") + 
  scale_y_continuous(expand = c(0.2, 0))
```

# Florian / RambowHYPERDiff

```{r}
data_dir
florian_sig_df <- read_tsv(file.path(data_dir, "signatures", "lista_signatures_florian_mod.tsv"))
hyperdiff_gs <- florian_sig_df$RambowHYPERdiff
hyperdiff_gs <- hyperdiff_gs[!is.na(hyperdiff_gs)]
```

```{r}
florian_df <- run_or_readRDS(
  file.path(file_cache_dir, "florian.rds"), 
  function() {
    gsva(
      tpm_matrix,
      gset.idx.list = list(RambowHYPERdiff = hyperdiff_gs),
      method = "ssgsea",
      kcdf = "Gaussian",
      min.sz = 3
    ) %>%
      as_tibble(rownames = "gs_name")
  }
)
florian_tidy <- florian_df %>% 
  pivot_longer(cols = 2:ncol(.), names_to = "sample_RNA", values_to = "score") %>% 
  dplyr::rename(NAME = gs_name) %>% 
  group_by(NAME) %>% 
  mutate(zscore = as.vector(scale(score))) %>% 
  ungroup()
```

```{r, fig.asp = 1/2, fig.width = 7}
florian_tidy %>% 
  left_join(bulk_clin_metadata, by = "sample_RNA") %>% 
  drop_na(response_splitSD) %>% 
  ggplot() + 
  aes(response_splitSD, score) +
  geom_boxplot(aes(fill = response_splitSD), 
               outlier.stroke = 0, 
               outlier.size = 0) + 
  geom_jitter(size = 2, shape = 1, width = 0.2) + 
  facet_grid(~treatment) + 
  stat_compare_means(comparisons = list(c("NR", "R"))) + 
  theme(legend.position = "none") + 
  remove_x_spine() + 
  scale_y_continuous(limits = ~c(..1[1], ..1[2]*1.1)) + 
  scale_fill_manual(values = palette_response) + 
  labs(x = "response", y = "ssGSEA score", title = "RambowHYPERdiff")
```

```{r, fig.asp = 1/2, fig.width = 7}
florian_tidy %>% 
  left_join(bulk_clin_metadata, by = "sample_RNA") %>% 
  drop_na(response) %>% 
  ggplot() + 
  aes(response, score) +
  geom_boxplot(aes(fill = response_splitSD), 
               outlier.stroke = 0, 
               outlier.size = 0) + 
  geom_jitter(size = 2, shape = 1, width = 0.2) + 
  facet_grid(~treatment) + 
  stat_compare_means(comparisons = list(c("NR", "R"))) + 
  theme(legend.position = "none") + 
  remove_x_spine() + 
  scale_y_continuous(limits = ~c(..1[1], ..1[2]*1.1)) + 
  scale_fill_manual(values = palette_response) + 
  labs(x = "response (drop all SD)", 
       y = "ssGSEA score", title = "RambowHYPERdiff")
```

```{r, fig.asp = 1/4, fig.width = 11}
florian_tidy %>% 
  left_join(bulk_clin_metadata, by = "sample_RNA") %>% 
  drop_na(response_splitSD) %>% 
  filter(treatment == "aPD1") %>% 
  ggplot() + 
  aes(response_splitSD, score) +
  geom_boxplot(aes(fill = response_splitSD), 
               outlier.stroke = 0, 
               outlier.size = 0) + 
  geom_jitter(size = 2, shape = 1, width = 0.2) + 
  facet_grid(~cohort) + 
  stat_compare_means(comparisons = list(c("NR", "R"))) + 
  theme(legend.position = "none") + 
  remove_x_spine() + 
  scale_y_continuous(limits = ~c(..1[1], ..1[2]*1.1)) + 
  scale_fill_manual(values = palette_response) + 
  labs(x = "response", 
       y = "ssGSEA score", title = "RambowHYPERdiff (aPD1 only)")
```

```{r, fig.asp = 1/4, fig.width = 11}
florian_tidy %>% 
  left_join(bulk_clin_metadata, by = "sample_RNA") %>% 
  drop_na(response) %>% 
  filter(treatment == "aPD1") %>% 
  ggplot() + 
  aes(response, score) +
  geom_boxplot(aes(fill = response), 
               outlier.stroke = 0, 
               outlier.size = 0) + 
  geom_jitter(size = 2, shape = 1, width = 0.2) + 
  facet_grid(~cohort) + 
  stat_compare_means(comparisons = list(c("NR", "R"))) + 
  theme(legend.position = "none") + 
  remove_x_spine() + 
  scale_y_continuous(limits = ~c(..1[1], ..1[2]*1.1)) + 
  scale_fill_manual(values = palette_response) + 
  labs(x = "response (drop all SD)", 
       y = "ssGSEA score", title = "RambowHYPERdiff (aPD1 only)")
```




